---
title: "Single cell preprocessing (AML + healthy)"
description: "not processed in this github"
date: "22 Apr 25 | 18 oct 25"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---

# sessionInfo of the versions of packages scrna is processed in

*** details: re-clustering using newer seurat for more efficient processing

- not ran with this current renv
```{plaintext}
R version 4.4.1 (2024-06-14 ucrt)
Platform: x86_64-w64-mingw32/x64
Running under: Windows 11 x64 (build 26200)

Matrix products: default


locale:
[1] LC_COLLATE=English_Singapore.utf8  LC_CTYPE=English_Singapore.utf8   
[3] LC_MONETARY=English_Singapore.utf8 LC_NUMERIC=C                      
[5] LC_TIME=English_Singapore.utf8    

time zone: Asia/Singapore
tzcode source: internal

attached base packages:
[1] grid      stats4    stats     graphics  grDevices datasets  utils     methods   base     

other attached packages:
 [1] harmony_1.0.1               Rcpp_1.1.0                  dendsort_0.3.4             
 [4] circlize_0.4.16             data.table_1.15.4           rstatix_0.7.2              
 [7] ggpubr_0.6.0                patchwork_1.3.2             RColorBrewer_1.1-3         
[10] ComplexHeatmap_2.20.0       SeuratExtend_1.2.5          SeuratExtendData_0.3.0     
[13] tradeSeq_1.18.0             Seurat_5.1.0                SeuratObject_5.0.2         
[16] sp_2.1-4                    lubridate_1.9.3             forcats_1.0.0              
[19] stringr_1.5.1               dplyr_1.1.4                 purrr_1.0.2                
[22] readr_2.1.5                 tidyr_1.3.1                 tibble_3.2.1               
[25] ggplot2_4.0.0               tidyverse_2.0.0             slingshot_2.12.0           
[28] princurve_2.1.6             TrajectoryUtils_1.12.0      SingleCellExperiment_1.26.0
[31] SummarizedExperiment_1.34.0 Biobase_2.64.0              GenomicRanges_1.56.1       
[34] GenomeInfoDb_1.40.1         IRanges_2.38.1              S4Vectors_0.42.1           
[37] BiocGenerics_0.50.0         MatrixGenerics_1.16.0       matrixStats_1.3.0          

loaded via a namespace (and not attached):
  [1] RcppAnnoy_0.0.22        splines_4.4.1           later_1.3.2            
  [4] polyclip_1.10-7         fastDummies_1.7.4       lifecycle_1.0.4        
  [7] doParallel_1.0.17       edgeR_4.2.2             globals_0.16.3         
 [10] lattice_0.22-6          pals_1.9                MASS_7.3-60.2          
 [13] backports_1.5.0         magrittr_2.0.3          sass_0.4.9             
 [16] limma_3.60.6            plotly_4.10.4           rmarkdown_2.28         
 [19] jquerylib_0.1.4         yaml_2.3.10             httpuv_1.6.15          
 [22] sctransform_0.4.1       spam_2.11-0             spatstat.sparse_3.1-0  
 [25] reticulate_1.38.0       cowplot_1.1.3           mapproj_1.2.11         
 [28] pbapply_1.7-2           maps_3.4.2.1            abind_1.4-5            
 [31] zlibbioc_1.50.0         Rtsne_0.17              WriteXLS_6.8.0         
 [34] GenomeInfoDbData_1.2.12 ggrepel_0.9.6           irlba_2.3.5.1          
 [37] listenv_0.9.1           spatstat.utils_3.0-5    goftest_1.2-3          
 [40] RSpectra_0.16-2         spatstat.random_3.3-1   fitdistrplus_1.2-1     
 [43] parallelly_1.38.0       leiden_0.4.3.1          codetools_0.2-20       
 [46] DelayedArray_0.30.1     tidyselect_1.2.1        shape_1.4.6.1          
 [49] UCSC.utils_1.0.0        farver_2.1.2            viridis_0.6.5          
 [52] spatstat.explore_3.3-1  jsonlite_1.8.8          GetoptLong_1.0.5       
 [55] progressr_0.14.0        iterators_1.0.14        ggridges_0.5.6         
 [58] survival_3.6-4          foreach_1.5.2           tools_4.4.1            
 [61] ica_1.0-3               glue_1.8.0              gridExtra_2.3          
 [64] SparseArray_1.4.8       mgcv_1.9-1              xfun_0.46              
 [67] withr_3.0.2             BiocManager_1.30.25     fastmap_1.2.0          
 [70] fansi_1.0.6             digest_0.6.36           timechange_0.3.0       
 [73] R6_2.5.1                mime_0.12               colorspace_2.1-1       
 [76] scattermore_1.2         tensor_1.5              dichromat_2.0-0.1      
 [79] spatstat.data_3.1-2     utf8_1.2.4              generics_0.1.3         
 [82] renv_1.0.11             httr_1.4.7              htmlwidgets_1.6.4      
 [85] S4Arrays_1.4.1          uwot_0.2.2              pkgconfig_2.0.3        
 [88] gtable_0.3.6            lmtest_0.9-40           S7_0.2.0               
 [91] XVector_0.44.0          htmltools_0.5.8.1       carData_3.0-5          
 [94] dotCall64_1.2           clue_0.3-65             scales_1.4.0           
 [97] png_0.1-8               spatstat.univar_3.0-0   enrichR_3.4            
[100] knitr_1.48              rstudioapi_0.16.0       tzdb_0.4.0             
[103] reshape2_1.4.4          rjson_0.2.21            nlme_3.1-164           
[106] curl_5.2.1              cachem_1.1.0            zoo_1.8-12             
[109] GlobalOptions_0.1.2     KernSmooth_2.23-24      parallel_4.4.1         
[112] miniUI_0.1.1.1          vipor_0.4.7             pillar_1.9.0           
[115] vctrs_0.6.5             RANN_2.6.1              promises_1.3.0         
[118] car_3.1-2               xtable_1.8-4            cluster_2.1.6          
[121] beeswarm_0.4.0          evaluate_0.24.0         locfit_1.5-9.12        
[124] cli_3.6.3               compiler_4.4.1          rlang_1.1.4            
[127] crayon_1.5.3            ggsignif_0.6.4          future.apply_1.11.2    
[130] plyr_1.8.9              ggbeeswarm_0.7.2        stringi_1.8.4          
[133] viridisLite_0.4.2       deldir_2.0-4            BiocParallel_1.38.0    
[136] lazyeval_0.2.2          spatstat.geom_3.3-2     Matrix_1.7-1           
[139] RcppHNSW_0.6.0          hms_1.1.3               future_1.34.0          
[142] statmod_1.5.0           shiny_1.9.1             ROCR_1.0-11            
[145] igraph_2.0.3            broom_1.0.6             bslib_0.8.0  
```



# loading libraries
```{r message=FALSE}
library(data.table)
library(tidyverse)
library(ggpubr)
library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(dendsort)
library(harmony)
library(RColorBrewer)

source("src/colors.R")
```

# processing all the cells as a whole

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE116256>
- patient samples AND healthy samples
- 38410 cells in total 

- separated the dataset into
  (): 1:35 malignant only 
  _w_healthy: 1:41 malignant with healthy
  _12aml: malignant selected for zeng's ppr (D0 with malignant cell population present)
  _h: healthy only


```{r}
data.dir = "data/"
count_files = list.files(data.dir, pattern = "*dem.txt", full.names = T)
anno_files = list.files(data.dir, pattern = "*anno.txt", full.names = T)
sample_name <- gsub("data/GSM.*_(.+)\\.dem.*", "\\1", count_files)

new_anno <- read.csv("scAML_new_annotations.csv") 
```

```{r}
merged_seurat = lapply(c(1:41), function(i){
  count_df <- read.csv(count_files[i], sep="\t", row.names = 1, check.names = F)
  anno_df <- read.csv(anno_files[i], sep="\t")
  anno_df <- anno_df[,!colnames(anno_df) %in% "NanoporeTranscripts"] # not present in every matrix
  anno_df <- left_join(anno_df, new_anno[,c(1,2,4)], by="Cell")
  rownames(anno_df) <- anno_df$Cell
  
  tmp <- CreateSeuratObject(counts = count_df, 
                     project = sample_name[i], 
                     meta.data = anno_df)
  
  Idents(object = tmp) <- factor(anno_df$CellType)
  tmp
})

saveRDS(merged_seurat, file = "output/all_samples/merged_seurat.rds")
# merged_seurat <- readRDS("output/all_samples/merged_seurat.rds")
```

# clustering and processing

(acc to the paper)
- removing n=8 samples from 4 low confidence samples - "AML314", "AML371", "AML722B", "AML997" 
- remaining n=29; N=12 + 5 healthy
- dropping cell cycle genes (n=10) 

```{r}
merged_seurat <- readRDS("output/all_samples/merged_seurat.rds")

all_ <- merge(x = merged_seurat[[1]], y = merged_seurat[2:41], merge.data = TRUE)

## combine to calculate MT
all_[["RNA"]] <- JoinLayers(all_[["RNA"]])
all_[["percent.mt"]] <- PercentageFeatureSet(all_, "^MT-")
all_@meta.data <- all_@meta.data %>%
  mutate(PatientID = gsub("-.*","", orig.ident))

all_ <- subset(all_, subset = PatientID %in% c("AML314", "AML371", "AML722B", "AML997"), invert = T) # low confidence and low count and poor quality samples

## split by patient
all_[["RNA"]] <- split(all_[["RNA"]], f = all_$PatientID)

all_ <- NormalizeData(all_, normalization.method = "LogNormalize", scale.factor = 10000)
all_ <- FindVariableFeatures(all_, selection.method = "vst", nfeatures = 3000)

VariableFeatures(object = all_) <- setdiff(VariableFeatures(object = all_),  c("ASPM", "CENPE", "CENPF", "DLGAP5", "MKI67", "NUSAP1", "PCLAF", "STMN1", "TOP2A", "TUBB") )

# all_ <- ScaleData(all_, vars.to.regress = "percent.mt") 
all_ <- ScaleData(all_) 
```


- batch effect
```{r}
all_ <- RunPCA(all_, features = VariableFeatures(object = all_), npcs = 50)

all_  <- FindNeighbors(all_, dims = 1:30, reduction = "pca")
all_  <- FindClusters(all_ , resolution = 0.6, cluster.name = "unintegrated_clusters")
all_ <- RunUMAP(all_, dims = 1:30, reduction = "pca", reduction.name = "umap.unintegrated")

DimPlot(all_, reduction = "umap.unintegrated", group.by = c("PatientID", "seurat_clusters")) + DimPlot(all_, reduction = "umap.unintegrated", group.by = c( "CellType"), cols =  alpha(hierarchy_colors,0.5))

ggsave("output/all_samples/plot/beforebc_all_2.png", width=20, height=5)
```


```{r warning=FALSE}
all_ <- IntegrateLayers(object = all_, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = 'harmony', npcs = 50)

all_ <- RunUMAP(all_, dims = 1:50, reduction = "harmony", n.neighbors = 30)

all_ <- FindNeighbors(all_, reduction = "harmony", dims = 1:50, k.param = 20)
all_ <- FindClusters(all_, resolution = 0.6)

DimPlot(all_, reduction = "umap", group.by = c("PatientID", "seurat_clusters"), alpha = 0.5) + DimPlot(all_, reduction = "umap", group.by = c( "CellType"), cols =  hierarchy_colors, alpha = 0.5)

ggsave("output/all_samples/plot/afterbc_all_2.png", width=20, height=5)

# saveRDS(all_, file="output/all_samples/all_.rds")
```

```{r}
FeaturePlot(all_, reduction = "umap", features = "percent.mt")

```

# saving r object

- Inflammation: The following features are not present in the object: IGLC1, IGHM, IGKC, IGLC2, IGHD, not searching for symbol synonyms
```{r}
all_ <- readRDS(file="output/all_samples/all_.rds")

all_[["RNA"]] <- JoinLayers(all_[["RNA"]])

all_$CellType <- factor(all_$CellType, levels=hierarchy_order[hierarchy_order %in% all_$CellType])
all_$celltype <- case_when(all_$CellType == 'HSC-like' ~ "HSC",
                              all_$CellType == 'Prog-like' ~ "Prog", 
                              all_$CellType == 'GMP-like' ~ "GMP", 
                              all_$CellType == 'ProMono-like' ~ "ProMono", 
                              all_$CellType == 'Mono-like' ~ "Mono", 
                              all_$CellType == 'cDC-like' ~ "cDC",
                              .default = all_$CellType)
all_$celltype <- factor(all_$celltype, levels=hierarchy_order[hierarchy_order %in% all_$celltype])

all_$cell_status <- paste0(ifelse(grepl("AML",all_$orig.ident) ,"AML","Healthy"), "_", all_$PredictionRefined)
all_$cell_status <- ifelse(all_$cell_status == "AML_normal", "AML_nonmalignant", all_$cell_status)

# to remove NA-783 & unclear -1133 (all belongs to BM5-34p38n) -> 36,494 cells
all_ <- subset(all_, cells = colnames(all_)[!all_$PredictionRefined %in% c("", "unclear")])

sampleinfo <- all_@meta.data 

# inflammation scores
reference <- read.csv("data/inflammatory_sig_246.csv")

all_ <- AddModuleScore(object = all_, features = list(reference$gene_id))
colnames(all_@meta.data)[length(colnames(all_@meta.data))] <- "Inflammation"

saveRDS(all_, file="~/data/scrna/all.rds")
```

## converting for compatibility of seurat v4
```{r}
all_[["RNA"]] <- as(object = all_[["RNA"]], Class = "Assay")

saveRDS(all_, file="~/data/scrna/all_v3.rds")
```

