---
title: "overview to ex vivo combination response data (fig 1/2, supp 1/2)"
description: "exploratory analysis of combi response"
date: "2 jan 26"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---

# libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rstatix)
library(data.table)

library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(dendsort)
library(grid)

library(ggpubr)
library(ggrepel)

set.seed(10)

source("src/function/colour.R")
source("src/function/functions.R")
source("src/function/loading_data.R")
```

# loading data
```{r}
matrix <- read.csv("data/oshu/sampleinfo_dss.csv", check.names = F)

mono <- read.csv("data/oshu/DSS_mono.csv") 

drugdet <- read.csv("data/oshu/mono_detail.csv")
listofdrug <- drugdet[!duplicated(drugdet$Drug),] %>%
  filter(! Drug %in% c("Arsenic Trioxide", "Quizartinib - NVP-TAE684"))

combi <- read.csv("data/oshu/DSS_combi.csv") 

combidet <- read.csv("data/oshu/combi_detail.csv")
listofcombi <- combidet[!duplicated(combidet$combo),]
```

# sample demographics

1. General summary of oshu data | Sample mapping

## suppfig1a: circos/samples overlapping between datasets

```{r}
oshu_SI <- matrix %>% 
  filter(Disease.status != "Healthy") %>%
  mutate(cohort = ifelse(grepl("S", Sample_ID), "BeatAML", "New")) %>% # sample membership
  select(Sample_ID, patientId, cohort) %>%
  arrange(patientId, cohort) %>%
  group_by(patientId) %>%
  mutate(sample_no = row_number(), 
         total_no = n(), # total no of samples
         type = n_distinct(cohort), # if the sample is in either cohort
         patient_existing = if_else(any(cohort == "BeatAML"), "BeatAML", "New") ) %>%
  ungroup() 

# distribution of samples in cohorts: type = 1 - unique to cohort, type = 2 - exist in both cohorts
oshu_SI %>% # sample level
  select(cohort, patient_existing, type) %>%
  table()

oshu_SI %>% # patient level
  distinct(patient_existing, patientId, type) %>%
  select(patient_existing, type) %>%
  table() 
```

, , type = 1

         patient_existing
cohort    BeatAML New
  BeatAML     325   0
  New           0 148

, , type = 2

         patient_existing
cohort    BeatAML New
  BeatAML      59   0
  New          64   0

                type
patient_existing   1   2
         BeatAML 303  49
         New     122   0
  

## Tbl 1: Sample Demographics

```{r}
drugcombo <- read.csv("data/oshu/oshu_sample_mapping.csv")
mut <- read.csv("data/oshu/mutation.csv")
top5 <- names(rev(sort(colSums(mut[,c(2:25,28:73)], na.rm = T))))[1:5]

df <- read.csv("data/oshu/clinical_summary.csv") |>
  left_join(mut[, c("Sample_ID", top5)]) |>
  select(gender, ageAtDiagnosis, FAB_subtype, ELN2022, consensusAMLFusions, Status, specimenType, outcome, all_of(top5), overallSurvival, vitalStatus) |>
  mutate(outcome = ifelse(is.na(outcome), "Unknown", outcome))


cbind(drugcombo,df)[1:596,] |>
  select(InBAML, hasRNA, hasDNA, hasProt, ageAtDiagnosis, gender, Status, ELN2022,  FAB_subtype, outcome, all_of(top5), overallSurvival, vitalStatus) |>
  gtsummary::tbl_summary(by=InBAML)
```


```{r}
cbind(drugcombo,df)[1:596,] |>
  select(InBAML, hasRNA, hasDNA, hasProt, ageAtDiagnosis, gender, Status, ELN2022,  FAB_subtype, outcome, all_of(top5), overallSurvival, vitalStatus) |> 
  gtsummary::tbl_summary() 
```

# overlapping data with existing cohort

Re-occurring drug screening data  + how many samples is involved
- monotherapy
- (included all counts, pre-filtering)

```{r}
oshu_sample <- read.csv("data/oshu/clinical_summary.csv")[,1:7]

baml <- read.csv("data/baml/mono_detail.csv")
ven <- read.csv("data/baml/ven_combi_detail.csv") %>% filter(Chemical_compound != "")

## from baml 2.0
dss_baml <- read.csv("data/baml/DSS_baml_indiv_raw.csv")[,c(3,26:28)] |>
  inner_join(oshu_sample[,c(2,5:7)]) |> # filter by sample
  left_join(baml[1:2],by=c("drug"="OGname")) |>
  filter(drug %in% unique(drugdet$Drug)) |> # filter by drugdet
  unique() |>
  select(Sample_ID, Chemical_compound) |>
  mutate(Tested = "BeatAML")

## from ven
dss_ven <- read.csv("data/baml/ven_combi_response.csv")[1:5] |>
  inner_join(oshu_sample[,c(2:4)], by=c("patient_id"="patientId", "lab_id"="labId")) |> # filter by sample
  inner_join(ven,by=c("drug")) |> # filter by combo
  unique() |>
  select(Sample_ID, Chemical_compound) 

check <- full_join(dss_baml, dss_ven) |>
  mutate(Tested = ifelse(is.na(Tested), "BeatAML-VenCombo", Tested))

## from oshu
dss_oshu <- read.csv("data/oshu/raw/DSS_oshu_raw.csv")[,c(3,26:27)] |>
  left_join(oshu_sample[,c(2:4)], by=c("patient"="patientId", "lab_id"="labId"))

dss_oshu_mono <- dss_oshu |>
  filter(!grepl(" - ", drug)) |>
  left_join(drugdet[1:2],by=c("drug"="Chemical_compound")) |>
  unique() |>
  select(Sample_ID, Drug) |>
  rename(Chemical_compound=Drug) |>
  mutate(Type = "monotherapy")

dss_oshu_combo <- dss_oshu |>
  filter(grepl(" - ", drug), drug != "Quizartinib - NVP-TAE684 - Arsenic Trioxide") |>
  left_join(combidet[3:4],by=c("drug"="Combination")) |> 
  unique() |>
  select(Sample_ID, combo) |>
  rename(Chemical_compound=combo) |>
  mutate(Type = "drug combination")
```

## Tbl 2: drug mapping - no of samples tested per drug per dataset
- based on what values exists/used in the current OSHU dataset

```{r}
dss_oshu <- rbind(dss_oshu_mono, dss_oshu_combo) |>
  left_join(check) |>
  mutate(Tested = ifelse(is.na(Tested), "BeatAML-DrugCombo", Tested)) |>
  group_by(Chemical_compound, Type, Tested) |>
  reframe(n=n()) |>
  pivot_wider(names_from = Tested, values_from = n, values_fill = list(n = NA)) |>
  arrange(Type, Chemical_compound)
dss_oshu$Total <- rowSums(dss_oshu[,3:5], na.rm = T)

write.csv(dss_oshu, "output/oshu_drug_mapping.csv", row.names = F)
```

```{r}
## new monotherapy in bamlcombi
sum(unique(drugdet$Drug) %in% unique(baml$Chemical_compound)) # 45
unique(drugdet$Drug) [!unique(drugdet$Drug) %in% unique(baml$Chemical_compound)]
```
- [1] "Arsenic Trioxide" "AGI combo"        "Telaglenastat"    "Dordaviprone"    
[5] "AZD5991"          "Decitabine" 

## SourceData: drug/combi detail - prefilter and postfilter counts

Total filtered counts considered
- number of samples left after QC and averaging
```{r}
# prefilter (unique samples)
dss_oshu_mono <- dss_oshu_mono |>
  summarise(Samples_tested = n(), .by = Chemical_compound) |>
  inner_join(listofdrug[2:8],by = c("Chemical_compound"="Drug")) |>
  arrange(tolower(Chemical_compound)) |>
  select(1,3:8,2)

dss_oshu_combo <- dss_oshu_combo |>
  left_join(listofcombi[c(1,2,4:5)], by = c("Chemical_compound"="combo")) |>
  group_by(Drug1, Drug2, Chemical_compound, combo_class) |>
  summarise(Samples_tested = n()) |>
  arrange(tolower(Chemical_compound))

# post filter (unique samples)
## sourcedata combi detail
dss_oshu_all <- read.csv("data/oshu/DSS_combi.csv") |>
  group_by(Combination) |>
  summarise(Samples_postfilt=n()) %>%
  left_join(dss_oshu_combo, ., by=c("Chemical_compound"="Combination"))

## sourcedata mono detail
dss_oshu_all <- read.csv("data/oshu/DSS_mono.csv") |>
  group_by(Chemical_compound) |>
  summarise(Samples_postfilt=n()) %>%
  left_join(dss_oshu_mono, .)
```

## suppfig 1b: correlation between mean DSS of newly added samples and prvs samples of the same drug
- those-not-exist-in-DrugCombo vs BeatAML_BeatAMLVen
- 48 monotherapy; 20 combinations
```{r}
# variable - listofdrug, mono, listofcombi, combi, oshu_sample, dss_ven

## from baml 2.0
dss_baml <- read.csv("data/baml/DSS_BAML_postfilt.csv") %>%
  filter(Chemical_compound %in% unique(listofdrug$Drug)) %>%
  mutate(label = paste0(Sample_ID, "_", Chemical_compound),
         Set = "BeatAML") 

## from ven
dss_ven <- dss_ven %>%
  mutate(label = paste0(Sample_ID, "_", Chemical_compound)) %>%
  mutate(Set = "BeatAMLVen")

############# mono ##############
dss_oshu <- mono %>%
  mutate(label = paste0(Sample_ID, "_", Chemical_compound)) %>%
  mutate(Set = ifelse(label %in% dss_baml$label, "BeatAML_BeatAMLVen", ifelse(label %in% dss_ven$label, "BeatAML_BeatAMLVen", "BeatAMLDrugCombo"))) %>%
  group_by(Chemical_compound, Set) %>%
  reframe(average_dss = mean(s_dss), n=n()) %>%
  group_by(Chemical_compound) %>%
  mutate(count=n())

# 48 common mono
df <- reshape2::dcast(dss_oshu[dss_oshu$count>1,], Chemical_compound ~ Set, value.var = "average_dss") %>%
  mutate(label="Monotherapy")

############# combi ##############
dss_oshu <- combi %>%
  rename(Chemical_compound = Combination) %>%
  mutate(label = paste0(Sample_ID, "_", Chemical_compound)) %>%
  mutate(Set = ifelse(label %in% dss_ven$label, "BeatAML_BeatAMLVen", "BeatAMLDrugCombo")) %>%
  group_by(Chemical_compound, Set) %>%
  reframe(average_dss = mean(s_dss), n=n()) %>%
  group_by(Chemical_compound) %>%
  mutate(count=n())

# 21 common combination
p <- reshape2::dcast(dss_oshu[dss_oshu$count>1,], Chemical_compound ~ Set, value.var = "average_dss") %>%
  mutate(label="Combination") %>%
  bind_rows(., df) %>%

  ggscatter(., x="BeatAML_BeatAMLVen", y="BeatAMLDrugCombo", 
            color = "label", palette = c("#a73030ff","#7aa6dcff"),
            add = c("reg.line"), add.params = list(color="grey20"),
            cor.coef = T, cor.method = "spearman"
            ) +
  theme(aspect.ratio = 1 ) +
  theme_pdf

ggsave(plot = p, "plot/suppfig1/supp_1b_scatterplot_cohort_drugresp.png", width=4, height = 4, units = "in", dpi = 600)
ggsave(plot = p, "plot/suppfig1/supp_1b_scatterplot_cohort_drugresp.pdf", width=4, height = 4, units = "in", bg = "transparent", device = cairo_pdf)
```

# fig 1b: heatmap overview
```{r}
# variable - matrix, listofcombi, combi

##### Heatmap for sampleinfo #####
oshu_SI <- matrix %>% 
  filter(Disease.status != "Healthy") %>%
  mutate(cohort = ifelse(grepl("S", Sample_ID), "BeatAML", "New"),
         AMLFusion = factor(outcome, levels=fusion.order),
         Disease.status = factor(Disease.status, levels=disease.order3, labels= c("Diagnosis", "Relapse", "Residual", "PostChemo", "Unknown")),
         ELN2022 = factor(ELN2022, levels=eln.col.order)) %>%
  arrange(Disease.status, ELN2022, FAB_subtype)

# Prepare mutation matrix
mat <- t(oshu_SI[,c(32:55, 58:103)])
top25 <- names(rev(sort(rowSums(mat, na.rm = T))))[1:25]
mat <- mat[top25,]
colnames(mat) <- oshu_SI$Sample_ID

na_mask <- data.frame(col = 1:596,
                      na = colSums(is.na(mat)),
                      count = colSums(mat, na.rm = T) ) %>%
  filter(na>0) %>%
  arrange(na, count)
na_mask <- na_mask$col

column_title = ""

col_ha = columnAnnotation(Cohort=oshu_SI$cohort,
  Disease_status = oshu_SI$Disease.status,
  FAB_subtype=oshu_SI$FAB_subtype,
  ELN2022=oshu_SI$ELN2022,
  AMLFusion=oshu_SI$AMLFusion,
  col=list(Cohort=c("BeatAML"="plum", "New"="turquoise4"),
           Disease_status = disease.col3,
           FAB_subtype=subtype.col2,
           ELN2022=eln.col,
           AMLFusion=fusion
           ),# na_col = "grey50",
  annotation_name_side = "left", height = unit(4, "cm"))

# Prepare matrix for oncoPrint - obtaining column order without NA first
mat2=mat
mat2[is.na(mat2)|mat2==0] = ""
mat2[mat2=="TRUE"] = "MUT;"
col = c("MUT" = "darkblue")
alter_fun = list(
    background = alter_graphic("rect", fill = "grey90", # "#CCCCCC", 
                               horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),  
    MUT = alter_graphic("rect", fill = col["MUT"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt"))
)

heatmap_legend_param = list(title = "Mutation", at = c("MUT"),  labels = c("Mutation"))

op <- oncoPrint(mat2, 
    alter_fun = alter_fun, col = col, 
    remove_empty_columns = FALSE, remove_empty_rows = TRUE,
    pct_side = "right", row_names_side = "left")

x <- ComplexHeatmap::draw(op)
col_dend = column_order(x)
col_dend = c(col_dend[!col_dend %in% na_mask], setNames(na_mask,na_mask))

# Oncoprint for mutation
mat[is.na(mat)] = "NA;"
mat[mat=="FALSE"] = "NEG;"
mat[mat=="TRUE"] = "MUT;"
col = c("MUT" = "darkblue",
        "NEG" = "grey90",
        "NA" = "grey50")
alter_fun = list(
    background = alter_graphic("rect", fill = "grey50", horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),   
    MUT = alter_graphic("rect", fill = col["MUT"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),
    NEG = alter_graphic("rect", fill = col["NEG"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),
    `NA` = alter_graphic("rect", fill = col["NA"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt"))
)

heatmap_legend_param = list(title = "Mutation", at = c("MUT", "NEG", "NA"), 
        labels = c("Present", "Absent", "NA"))

op <- oncoPrint(mat, 
    alter_fun = alter_fun, col = col, column_order = col_dend, 
    remove_empty_columns = FALSE, remove_empty_rows = TRUE,
    pct_side = "right", row_names_side = "left", top_annotation = col_ha,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    bottom_annotation = HeatmapAnnotation(cbar = anno_oncoprint_barplot(type = c("MUT"),axis_param = list(direction="reverse"))),
    right_annotation = rowAnnotation(rbar = anno_oncoprint_barplot(type = c("MUT"))),
    pct_include = c("MUT"),
    height = unit(9, "cm")
    )

# Drug response heatmap
mat <- reshape2::acast(combi, Combination~Sample_ID, value.var = "s_dss", fun.aggregate = mean, fill = NA_real_)
mat <- mat[order(-rowSums(!is.na(mat))), ]

mask <- !is.na(mat)
row_anno <- listofcombi[order(listofcombi$combo, rownames(mat)),4:5] %>%
  separate_wider_delim(cols=combo_class, delim=" - ", names = c("class1", "class2"), cols_remove = T)

row_anno = rowAnnotation(
  `Drug class 1` = row_anno$class1,
  `Drug class 2` = row_anno$class2,
  col=list(`Drug class 1` = drugclass, `Drug class 2` = drugclass),
  gp = gpar(col = "grey50", lwd=.1), 
  annotation_name_side = "top",
  annotation_name_rot=45,
  annotation_legend_param = list( 
    `Drug class 1` = list(
      title="Drug Class", 
      at=c("Apoptotic modulator", "Chemotherapeutics", "Differentiating/epigenetic modifier", "Immunomodulator", "Kinase inhibitor", "Metabolic modifier", "mTOR inhibitor", "Other"), 
      legend_gp = gpar(fill = c("#E41A1C", "#FFD92F", "#4DAF4A", "#FF7F00", "royalblue", "darkolivegreen1", "turquoise4", "plum")) 
      )
    ),
  show_legend = c(TRUE, FALSE)
)


dr <- Heatmap(
  mat, na_col = "grey90", rect_gp = gpar(col = "white", lwd = 0.2),
  col= colorRamp2(seq(0,1,by=0.1), viridis::viridis(n=11)),
  row_title = "Combinations",
  column_title = "Samples",
  show_row_names = F, show_column_names = F,
  show_row_dend = F, show_column_dend = F,
  cluster_rows = F, cluster_columns = F, column_order = col_dend,
  right_annotation = row_anno,
  heatmap_legend_param = list(legend_height = unit(8, "cm"), direction = "horizontal"),
  name = "DSS", height = unit(10, "cm")
)

ht_list = dr %v% op 

Cairo::CairoPDF("plot/fig1/main_1b_heatmap_overview_demo.pdf", width=16.5, height=12.5, bg = "transparent", family = "Arial")
# png("plot/fig1/main_1b_heatmap_overview_demo.png", width=16.5, height=12.5, units="in", res=600, family = "Arial", bg = "transparent")
ComplexHeatmap::draw(ht_list, heatmap_legend_side = "bottom", annotation_legend_side = "bottom", merge_legend=T, background = "transparent")
dev.off()
```

# suppfig 2a: dss distribution of indiv monotherapy
```{r}
# variable - mono, listofdrug
plot <- mono %>%
  group_by(Chemical_compound) %>%
  mutate(median = median(s_dss), n=n(),
         label = paste0(Chemical_compound, " (n=", n, ")")) %>%
  left_join(., unique(listofdrug[,c(2,7)]), by=c("Chemical_compound"="Drug"))

p <- ggplot(plot, aes(x=reorder(label, median, decreasing = T), y=s_dss)) +
  geom_violin(aes(colour=Drug_Class), scale="width", linewidth=1) + 
  ggforce::geom_sina(size=0.2, scale="width", colour="black") +
  scale_colour_manual(values = drugclass) +
  labs(x="", y="DSS") +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1),
        legend.position = "right", legend.text = element_text(size=11),
        legend.key.height = unit(0.5,"cm")
        ) + 
  theme_pdf

ggsave(plot = p, "plot/suppfig2/supp_2a_violinplot_mono_dss.png", width = 20, height = 5, units = "in", dpi=600)
ggsave(plot = p, "plot/suppfig2/supp_2a_violinplot_mono_dss.pdf", width = 20, height = 5, units = "in", bg = "transparent", device = cairo_pdf)
```

# fig 2a: dss distribution
```{r}
# variable - mono, combi

## Prepare data for plotting - combining and converting to long format
p <- mutate(combi[,c(5,8)], Values="Combination") %>%
  dplyr::rename(Chemical_compound = Combination) %>%
  rbind(., mutate(mono[,c(3,6)], Values="Monotherapy")) %>%
  mutate(Values = factor(Values, levels=c("Combination", "Monotherapy")))

# Wilcoxon test p-value
pval <- compare_means(s_dss ~ Values, data = p, method = "wilcox.test") # 0

plot <- ggplot(p , aes(y=Values,x=s_dss, fill=Values)) +
  scale_fill_manual(values=c("#a73030ff", "#7aa6dcff"), guide="none") +
  ggridges::geom_density_ridges(quantile_lines=TRUE, scale = 1, rel_min_height = 0.01, alpha=0.7, quantiles=2)  + 
  labs(x="DSS", y="") +
  theme_pubr(base_size = 14) +
  theme(aspect.ratio = 0.5,
        panel.grid.major.y = element_line(color = "#A8BAC4", linewidth = 0.3) ) + 
  annotate("text", x = max(p$s_dss) * 0.95, y = 2.8, label = "p < 2.2e-16", hjust = 1, size = 4)

plot <- plot + 
  theme_pdf

ggsave(plot = plot, "plot/fig2/main_2a_ridgeplot_overalldss.png", width = 6, height = 3, units = "in", dpi=600)
ggsave(plot = plot, "plot/fig2/main_2a_ridgeplot_overalldss.pdf", width = 6, height = 3, units = "in", bg = "transparent", device = cairo_pdf)
```

# tbl 3. correlation between monotherapy and combination therapy

```{r}
sampleinfo <- matrix[1:596,]

mono.m <- sampleinfo[,c(1,104:154)] %>% column_to_rownames("Sample_ID") %>%
  select(-`Arsenic Trioxide`)
combi.m <- sampleinfo[,c(1,155:324)] %>% column_to_rownames("Sample_ID")

df <- bind_cols(mono.m, combi.m)

temp <- .run.cor(group.m=df, x.m=df, min.n = 5, method = "spearman") %>%
  rowwise() %>%
  mutate(Category_1 = ifelse(grepl(" - ", Group), "drug combination", "monotherapy"),
         Category_2 = ifelse(grepl(" - ", Combination), "drug combination", "monotherapy")
         )

write.csv(temp, "output/oshu_drug_sens_correlation.csv", row.names = F)
```

## suppfig 2b: for correlation between monotherapy response (table 3)
```{r}
listofdrug1 <- listofdrug %>% arrange(Drug_Class, Drug_Sub_Class, MOA) %>%
  arrange(Drug_Class, Drug_Sub_Class) %>%
  mutate(label = str_trunc(Drug, width = 13, ellipsis = ""))

temp <- read.csv("output/oshu_drug_sens_correlation.csv") %>%
  mutate(Group = str_trunc(Group, width = 13, ellipsis = ""),
         Combination = str_trunc(Combination, width = 13, ellipsis = "")) %>%
  filter(Category_1 == "monotherapy", Category_2 == "monotherapy", cor != 1)
temp <- reshape2::acast(Group~Combination, data=temp, value.var = "cor")
temp <- temp[listofdrug1$label, listofdrug1$label]

## plotting heatmap
row_anno = rowAnnotation(
  `DrugClass` = listofdrug1$Drug_Class,
  col=list(`DrugClass` = drugclass),
  gp = gpar(col = "white", lwd=.1), 
  annotation_name_side = "top",
  annotation_name_rot=0,
  annotation_legend_param = list( 
    `DrugClass` = list(
      title="Drug Class", 
      at=c("Apoptotic modulator", "Chemotherapeutics", "Differentiating/epigenetic modifier", "Immunomodulator", "Kinase inhibitor", "Metabolic modifier", "mTOR inhibitor", "Other"), 
      legend_gp = gpar(fill = c("#E41A1C", "#FFD92F", "#4DAF4A", "#FF7F00", "royalblue", "darkolivegreen1", "turquoise4", "plum")) 
      )
    )
)

hm <- Heatmap(
  temp, rect_gp = gpar(type = "none"), na_col = "white",
  colorRamp2(c(-0.6, -0.3, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.75, 1), 
             rev(brewer.pal(10, "Spectral"))),
  cluster_rows = F, cluster_columns = F,
  show_row_dend = F, show_column_dend = F, left_annotation = row_anno,
  row_names_side = "left", column_names_rot = 50,
  heatmap_width = unit(11, "in"), heatmap_height = unit(10, "in"),
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(i >= j) { grid.rect(x, y, w, h, gp = gpar(fill = fill, col = "grey"))}
    if(i==j) {grid.rect(x, y, w, h, gp = gpar(fill = "white", col = "grey"))}
  },
  name = "Spearman corr"
  )

png("plot/suppfig2/supp_2b_heatmap_corr_mono.png", width = 14, height = 11, units = "in", res = 600, family = "Arial", bg = "transparent")
# Cairo::CairoPDF("plot/suppfig2/supp_2b_heatmap_corr_mono.pdf", width=14, height = 11, bg = "transparent", family = "Arial")
ComplexHeatmap::draw(hm, background = "transparent")
dev.off()
```

# combination ratio/index of each combination

```{r}
## calculating CR at samples levels using HSA/IDA prinicple
HSA_eval <- rbindlist(
  lapply(setNames(1:170, listofcombi$combo), function(i){
    
    # combination DSS per sample
    data <- combi[combi$Combination==listofcombi[i, 4],] %>%
      dplyr::select(Sample_ID, Combination, s_dss)
    
    # monotherapy DSS per sample
    data <- mono[mono$Chemical_compound %in% listofcombi[i, 1:2],] %>%
      reshape2::dcast(Sample_ID ~ Chemical_compound, fun.aggregate = mean, value.var = "s_dss", fill = NA_real_) %>%
      inner_join(., data, by="Sample_ID")
    
    data <- data[complete.cases(data),]
    colnames(data)[2:3] <- c("Drug1_dss", "Drug2_dss")
    
    data$best_mono <- apply(data[,2:3], 1, function(x){ifelse(x[1]==x[2], NA, listofcombi[i, which.max(x)])})
    data$HSA_ref = pmax(data$Drug1_dss, data$Drug2_dss)
    data$Excess_HSA = data$s_dss - data$HSA_ref
    data$CR <- data$HSA_ref/data$s_dss # predicted/observed
    data$CR[is.nan(data$CR)] = NA # Inf when combi=0; NaN when all is 0
    data
  }) 
)

write.csv(HSA_eval, "output/supp/oshu_HSA_eval.csv", row.names = F)
```

## tbl 4: combination index dss
```{r}
## selecting for correlations of combi with constituent mono
temp <- read.csv("output/oshu_drug_sens_correlation.csv") %>%
  filter(Category_1 == "monotherapy", Category_2 == "drug combination") %>%
  rowwise() %>%
  mutate(
    cor = signif(cor, 3),
    select = grepl(Group, Combination),
    drug_partner = if_else( select, gsub(" - ", "", gsub(Group, "", Combination)), NA_character_ )
  ) %>%
  filter(select) %>% ungroup()

## summarising across samples for each combinations
HSA_eval <- read.csv("output/supp/oshu_HSA_eval.csv", check.names = F) %>%
  group_by(Combination) %>%
  summarise(Median_Drug1_dss = median(Drug1_dss), 
            Median_Drug2_dss = median(Drug2_dss), 
            Median_dss = median(s_dss), 
            median_CR = median(CR, na.rm=T), n = n(), 
            percentSample = sum(CR<1 , na.rm = T)*100/n )

## adding in the correlation info to the constituent mono & combination type
combiclass <- combidet %>% select(4,5)  %>% distinct() %>% column_to_rownames("combo")

temp <- temp %>%
  arrange(Group) %>%
  group_by(Combination) %>%
  mutate(Drug = paste0("Drug", row_number())) %>%
  select(Drug, Combination, cor) %>%
  pivot_wider(names_from = Drug, values_from = cor) %>%
  rename(Drug1_corr = Drug1, Drug2_corr = Drug2)

temp <- left_join(HSA_eval, temp) %>%
  mutate(
    One_driver = (Drug1_corr > 0.5|Drug2_corr > 0.5)& abs(Drug1_corr - Drug2_corr)>0.1,
    One_driver = ifelse(One_driver, ifelse(Drug1_corr > Drug2_corr, "Drug1", "Drug2"), "None"),
    Better_combi = median_CR < 1,
    Combo_class = combiclass[Combination,]
    ) %>%
  select(1, 12, 2:11) %>% 
  arrange(median_CR)

write.csv(temp, "output/oshu_combinationindex.csv", row.names = F)
```

## fig 2b: scatterplot median CR vs median DSS
```{r}
# variable - HSA_df
HSA_df <- read.csv("output/supp/oshu_HSA_eval.csv", check.names = F) %>%
  reframe(median_dss = median(s_dss, na.rm=T), 
          n = n(), 
          median_CR = median(CR, na.rm=T), 
          median_excesshsa = median(Excess_HSA, na.rm=T),
          fill = sum(CR<1 , na.rm = T)*100/n, .by=c(Combination)) 

plot <- HSA_df %>%
  ungroup() %>%
  mutate(
    median_CR = if_else(median_CR > 3, 3.3, median_CR),
    label_up = if_else((median_CR > 1.5 | median_dss > 0.46) & median_CR <= 3 & median_CR > 1, Combination, NA_character_),
    label_down = if_else(median_CR < 1 & median_dss > 0.46, Combination, NA_character_),
    label2 = if_else(Combination == "Azacitidine - Venetoclax", "Azacitidine - Venetoclax", NA_character_)
  )

ticklabels <- c("0.2", "1", ">3")
tickvals <- c(0.2, 1, 3)

plot <- plot %>%
  ggplot(., aes(x=median_dss, y=median_CR, size=n, fill=fill)) +
  geom_point(color="black", alpha=0.5, shape=21) + 
  geom_hline(yintercept = 1, linetype="dashed") +
  scale_y_log10(labels = ticklabels, breaks = tickvals, limits = c(0.2,3.5)) +
  
  geom_text_repel(mapping=aes(label=label_up), size=4, xlim = c(0,Inf), ylim = c(log10(1), log10(3)),
                  min.segment.length = 0, box.padding=.5, point.padding=.25,
                  max.overlaps = 9, force = 1,
                  seed=123) +

  geom_text_repel(mapping=aes(label=label_down), size=4, xlim = c(0,1.25), 
                  ylim = c(log10(0.2), log10(1)),
                  min.segment.length = 0, box.padding=.5, point.padding=.25,
                  max.overlaps = 20, force = 0.2,
                  seed=123) +
  
  geom_text_repel(mapping=aes(label=label2), point.padding = unit(0.5, "lines"),
              segment.color = "black", min.segment.length = 0.03, 
              color = "blue", size=4.5, fontface="bold", nudge_x = -0.4, nudge_y = -0.05) + 
  
  scale_fill_viridis_c(name="% Samples\nwith CR < 1", limits=c(0,100)) +
  scale_size_continuous(name="Sample size") +
  
  annotate( "text", label = "values above this threshold are capped", x=1, y=3.2, colour = "grey50" ) +
  geom_hline(yintercept = 3, linetype="dashed") +
  xlim(c(0,1.25)) + 
  labs(x = "Median DSS", y = "Median Combination Index") +
  theme_pubr(base_size = 14) +
  theme(aspect.ratio = 0.9,
        legend.title = element_text(size=12), 
        legend.text = element_text(size = 12),
        legend.position = "top")

plot <- plot + 
  theme_pdf

ggsave(plot = plot, "plot/fig2/main_2c_scatterplot_CR.png", units = "cm", width = 23, height = 21, dpi=600)
ggsave(plot = plot, "plot/fig2/main_2c_scatterplot_CR.pdf", units = "cm", width = 23, height = 21, bg = "transparent", device = cairo_pdf)
```

## fig 2c: circos figure 2 for table 4
```{r}
## use HSA_df
df <- read.csv("output/oshu_combinationindex.csv") %>%
  mutate(Better_combi = case_when(median_CR == 1 ~ "CR=1",
                                  median_CR < 1 ~ "CR<1",
                                  median_CR > 1 ~ "CR>1"))

combi <- read.csv("data/oshu/DSS_combi.csv") 
plot <- combi %>%
  group_by(Combination) %>%
  reframe(
    pt25 = quantile(s_dss,probs=.25),
    median = median(s_dss),
    pt75 = quantile(s_dss, probs=.75),
    n=n()
  ) %>%
  mutate(label = paste0(" (n=", n, ")")) %>%
  left_join(., unique(listofcombi[,c(1,2,4,5)]), by=c("Combination"="combo")) %>%
  left_join(., unique(listofdrug[,c(2,7)]), by=c("Drug1"="Drug")) %>%
  left_join(., unique(listofdrug[,c(2,7)]), by=c("Drug2"="Drug"), suffix = c("1", "2"))  %>%
  mutate(
    drug1_colour = drugclass[Drug_Class1], 
    drug2_colour = drugclass[Drug_Class2]
  ) %>% column_to_rownames("Combination")

combi <- reshape2::acast(Sample_ID~Combination, data=combi, value.var = "s_dss")

drugclass_ = c("Apoptotic\nmodulator" = "#E41A1C", 
              "Chemotherapeutics" ="#FFD92F",
              "Differentiating\nepigenetic modifier" = "#4DAF4A",
              "Immunomodulator" ="#FF7F00",
              "Kinase inhibitor" = "royalblue",
              "Metabolic modifier" = "darkolivegreen2",
              "mTOR inhibitor" = "turquoise4",
              "Other" = "plum"
)

```


```{r}
# Order rows 
df_ord <- df %>% column_to_rownames("Combination") %>%
  arrange(Median_dss)

mat_ord <- as.matrix(df_ord %>% select(Drug1_corr, Drug2_corr) )
combi_ord <- combi[,rownames(df_ord)]
plot_ord <- plot[rownames(df_ord),]

# Colors
col_corr <- colorRamp2(
  c(-1, -0.75, 0, 0.75, 1),
  c("#4393c3","#2b8cbe", "white", "#e34a33", "#d6604d") )

col_dss  <- colorRamp2(
  c(0, 0.3, 1),
  c("#f7fcf5", "#a1d99b", "seagreen") )
  
# Draw the 2-column tiny heatmap around the circle
# png("plot/fig2/main_2c_circos.png", width = 15, height = 15, units = "in", res = 600, bg = "transparent", family = "Arial")
Cairo::CairoPDF("plot/fig2/main_2b_circos.pdf", width = 15, height = 15, bg = "transparent", family = "Arial")

circos.clear()
circos.par(gap.after = c(20),
           start.degree = 80,
           cell.padding = c(0, 0, 0, 0),
           track.margin = c(0.005, 0.001),
           circle.margin = c(0.001, 0.001, 0.001, 0.001) 
)

circos.heatmap.initialize(mat_ord, cluster = FALSE)

##  combination names
part1 <- str_trunc(plot_ord$Drug1, width = 13, ellipsis = "")          # length 170
part2 <- rep("-", 170)
part3 <- str_trunc(plot_ord$Drug2, width = 13, ellipsis = "")          # length 170
part4 <- plot_ord$label
col1  <- plot_ord$drug1_colour   # length 170
col2  <- rep("black", 170)
col3  <- plot_ord$drug2_colour   # length 170

circos.trackPlotRegion(
  ylim = c(0, 1), track.height = 0.35, bg.border = NA,
  panel.fun = function(x, y) {
    pos  <- seq(0.5, 169.5, by = 1)   # 1:170
    y0   <- 0                         # middle of the track (radial)

    # draw three fragments per slot, horizontally separated
    circos.text(pos, y0, labels = part1, col = col1,
                facing = "clockwise", niceFacing = TRUE, cex = 1, adj = c(0, 0.5))
    circos.text(pos, y0 + 0.47, labels = part2, col = col2,
                facing = "clockwise", niceFacing = TRUE, cex = 1, adj = c(0.5, 0.5))
    circos.text(pos, y0 + 0.5, labels = part3, col = col3,
                facing = "clockwise", niceFacing = TRUE, cex = 1, adj = c(0, 0.5))
    circos.text(pos, y0 + 0.97, labels = part4, col = col2,
                facing = "clockwise", niceFacing = TRUE, cex = 1, adj = c(0, 0.5))
  }
)

circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  circos.text(CELL_META$cell.xlim[2] + convert_x(25, "mm"), y = 0.55, 
              labels = "Sample no.\n\n\n\n\n\n\nDrug2\n\n\n\n\n\n\nDrug1", cex = 0.8, adj = c(0.5, 0.5))
  }, bg.border = NA)

## correlation
circos.heatmap(
  mat_ord[,2:1], col = col_corr, cluster = FALSE,
  # rownames.side = "outside", rownames.cex = 1,
  show.sector.labels = FALSE,
  track.height = 0.1, cell.border = "white", cell.lwd = .5
)

## correlation text
circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  if(CELL_META$sector.numeric.index == 1) { # the last sector
    cn = c("cor(Drug1,Combi)", "cor(Drug2,Combi)")
    n = length(cn)
    circos.text(x = rep(CELL_META$cell.xlim[2], n) + convert_x(17, "mm"), y = 1:n - 0.5, 
                labels = cn, cex = .8, adj = c(0.5, 0.5), facing = "bending.inside")
  }
}, bg.border = NA)

## boxplot/pointrange
circos.track(ylim = c(0, 1), panel.fun = function(x, y) {
  y = seq(0.25, 0.75, by = 0.25)
  circos.segments(0, y, 170, y, col = "grey50", lty = 3)
  circos.segments(1:169, 0.02, 1:169, 0.98, col = "grey80", lty = 3)
  
  for(pos in seq(1, 170, by = 1)) {
    x=pos-0.5
    value = plot_ord[pos,]
    circos.segments(x, value$pt25, x, value$pt75)
    
    cap <- mm_x(0.8)
    circos.segments(x - cap, value$pt25, x + cap, value$pt25)
    circos.segments(x - cap, value$pt75, x + cap, value$pt75)
    circos.points(x, value$median, pch = 16, cex = 0.5)
    
    # value = combi_ord[,pos] %>% na.omit() %>% as.numeric()
    # circos.boxplot(value, pos-0.5, outline = F, box_width = 0.5, cex = .7)
  }
  }, track.height = 0.1
)
circos.yaxis(side = "right", labels.cex = .7, at = seq(0, 1, by = 0.25))

circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  circos.text(CELL_META$cell.xlim[2] + convert_x(16, "mm"), y = 0.5, 
              labels = "DSS\n(25th,75th)", cex = 0.8, adj = c(0.5, 0.5))
  }, bg.border = NA)

## median dss
circos.heatmap(as.numeric(df_ord[[4]]) , col = col_dss, cluster = FALSE, track.height = 0.02, cell.border = "white", cell.lwd = .5 )

circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  circos.text(CELL_META$cell.xlim[2] + convert_x(12,"mm"), y = 0.5, 
              labels = "MedianDSS", cex = 0.8, adj = c(0.5, 0.5), facing = "bending.inside")
  }, bg.border = NA)

## one driver, better combi
circos.heatmap(as.character(df_ord[[11]]) , col = c(`CR<1` = "#3b5b92", `CR=1` = "darkturquoise", `CR>1` = "lavender"), cluster = FALSE, track.height = 0.03, cell.border = "white", cell.lwd = .5)

circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  circos.text(CELL_META$cell.xlim[2] + convert_x(11.5, "mm"), y = 0.5, 
              labels = "Median CR", cex = 0.8, adj = c(0.5, 0.5), facing = "bending.inside")
  }, bg.border = NA)

circos.heatmap(as.character(df_ord[[10]]) , col = c(Drug1 = "#FC8D59", None = "lemonchiffon", Drug2 = "burlywood3"), cluster = FALSE, track.height = 0.03, cell.border = "white", cell.lwd = .5 )

circos.track(track.index = get.current.track.index(), panel.fun = function(x, y) {
  circos.text(CELL_META$cell.xlim[2] + convert_x(10, "mm"), y = 0.5, 
              labels = "OneDriver", cex = 0.8, adj = c(0.5, 0.5), facing = "bending.inside")
  }, bg.border = NA)

# dev.off()

# png("plot/circos_legend_table4.png", width = 10, height = 10, units = "in", res = 600)  # High-res, large file
# circos.clear()
lgd_corr <- Legend(title = "Spearman corr", col_fun = col_corr, legend_height = unit(1.5, "cm"))
lgd_dss  <- Legend(title = "Median DSS", at = pretty(c(0, 1), 2), col_fun = col_dss, legend_height = unit(1.5, "cm"))
lgd_better = Legend(title = "Median CR", at = c("CR<1", "CR=1", "CR>1"), 
    legend_gp = gpar(fill = c("#3b5b92", "darkturquoise", "lavender"))) 
lgd_driver = Legend(title = "OneDriver", at = c("Drug1", "None", "Drug2"), 
    legend_gp = gpar(fill = c("#FC8D59","lemonchiffon","burlywood3")))
lgd_dc = Legend(title = "Drug Class", at = names(drugclass_), 
    legend_gp = gpar(fill = drugclass_))

draw(packLegend(lgd_corr, lgd_dss, lgd_better, lgd_driver, lgd_dc, max_height = unit(8, "cm")))

dev.off()
```

## fig 2d: barplot for frequency occurrence of drugs
- originially created in google sheets
```{r}
# variable - HSA_df, combi
HSA_df <- read.csv("output/oshu_combinationindex.csv")

tmp <- listofcombi %>% dplyr::select(Drug1, Drug2) %>% ## freq of occurrence for each drug
  pivot_longer(cols = 1:2,values_to = "Drug") %>%
  reframe(n = n(), .by = "Drug")
tmp <- setNames(tmp$n, tmp$Drug)

selected <- HSA_df %>% 
  filter(median_CR < 1 & Median_dss > 0.426) %>%
  left_join(., listofcombi[,c(1,2,4)], by=c("Combination"="combo")) %>%
  dplyr::select(Drug1, Drug2) %>%
  pivot_longer(cols = 1:2,values_to = "Drug") %>%
  reframe(n = n(), Proportion = round(n/68, 2), .by = "Drug") %>%
  mutate(total = tmp[Drug], normalised_prop = ifelse(total<5, NA, round(n/total,2)) ) %>%
  left_join(., listofdrug[,c(2,5:8)]) %>% 
  slice_max(n=10, order_by = normalised_prop) %>%
  mutate( Approval_status = ifelse(Approval_status == "Approved", "Approved", "Investigational") ) 

selected$label <- c( "Birabresib (BETi)", "Birinapant (IAPi)", "AZD5991 (MCL1i)", "Trametinib (MEKi)", "Palbociclib (CDKi)", "Idelalisib (PI3Ki)", "Bortezomib (Proteasome-i)", "JQ1 (BETi)", "Venetoclax (BCL2i)", "Quizartinib (FLT3i)" )


ggplot(selected, aes(x = normalised_prop, y = reorder(label, normalised_prop, decreasing = F), fill = Approval_status)) +
  geom_bar(stat = "identity", color = NA, width = 0.6) +
  scale_fill_manual(values = c("#fe9900", "#0f518e"), name = "") +
  geom_text(aes(label = round(normalised_prop, 2)), hjust = 1.1, color = "white") +
  scale_x_continuous(expand = c(0, 0), limits = c(0,1)) +
  xlab("Frequency of occurrence (normalised)") + ylab("") +
  theme_pubr(base_size = 12) +
  theme(
    legend.key.height = unit(0.5, "cm"),
    aspect.ratio = 0.8,
    panel.grid.major.x = element_line(color = "#cccccc")
  ) +
  theme_pdf

ggsave("plot/fig2/main_2d_barplot_topapproved.png", width = 6, height = 4, units = "in", dpi=600, bg = "transparent")
ggsave("plot/fig2/main_2d_barplot_topapproved.pdf", width = 6, height = 4, units = "in", bg = "transparent", device = cairo_pdf)
```

## suppfig 2c - created using STITCH
<http://stitch.embl.de/>

## fig 2e - boxplot for clinically approved partners
```{r}
HSA_eval <- read.csv("output/supp/oshu_HSA_eval.csv", check.names = F)

df <- read.csv("output/oshu_combinationindex.csv") %>%
  left_join(listofcombi[,c(1,2,4)], by=c("Combination"="combo")) %>%
  filter(median_CR <= 1, Median_dss >= 0.426)

selected_approved <- c("Venetoclax", "Quizartinib", "Sorafenib", "Palbociclib", "Idelalisib", "Trametinib", "Bortezomib") # in 44 unique combi

df_long <- df %>% select(Combination,Drug1,Drug2) %>% reshape2::melt(id.vars = "Combination", value.name = "Drug") %>%
  filter(Drug %in% selected_approved)

plotlist <- list()
heights <- c()
for (i in selected_approved) {
  df2 <- df_long %>% filter(Drug == i) %>%
    distinct(Combination) %>%
    left_join(df) %>%
    arrange(Median_dss)
  
  df2_long <- HSA_eval %>%
    filter(Combination %in% df2$Combination) %>%
    group_by(Combination) %>%
    mutate(fill = sum(CR<1 , na.rm = T)*100/n())
  
  plotlist[[i]] <- df2_long %>%
    mutate(Combination = factor(Combination, levels=df2$Combination)) %>%
    ggplot(., aes(x=Combination, y=s_dss, fill = fill)) +
    geom_boxplot(width = 0.5, color = "black") +
    viridis::scale_fill_viridis(name="% Samples\nwith CR < 1", limits=c(45,92)) +
    labs(x="", y="DSS", title = i) +
    coord_flip() +
    theme_pubr() +
    theme(panel.grid.major.x = element_line(color = "#A8BAC4", size = 0.3),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "right") 
    
  heights <- c(heights, nrow(df2)/61)
}

p <- patchwork::wrap_plots(plotlist, ncol = 1, heights = heights, guides = "collect") &
    theme_pdf


ggsave(plot = p, "plot/fig2/main_2e_boxplot_clinical.png", width = 6, height = 20, dpi = 600, bg = "transparent")
ggsave(plot = p, "plot/fig2/main_2e_boxplot_clinical.pdf", width = 6, height = 20, bg = "transparent", device = cairo_pdf)

```

# sessioninfo
```{r}
sessionInfo()
```

