---
title: "IDA prediction ()"
description: "Computational predictions of combination responses"
date: "2 jan 26"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---

# loading libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rstatix)
library(data.table)
library(ggpubr)
library(ggrepel)

set.seed(10)
source("src/function/plot.R")
source("src/function/colour.R")
source("src/function/functions.R")
```

# 5. IDAPrediction

```{r}
source("src/function/ida_valid.R")
listofcombi <- combi <- read.csv("data/oshu/combi_detail.csv") %>% select(1,2,4)  %>% distinct() 

mono <- read.csv("data/oshu/DSS_mono.csv") %>%
  dplyr::select(Sample_ID, Chemical_compound, Concentration, s_dss) %>%
  arrange(Sample_ID, Chemical_compound)

combi <- read.csv("data/oshu/DSS_combi.csv") %>%
  dplyr::select(Sample_ID, Drug1, Drug2, Combination, Concentration, s_dss) %>%
  dplyr::rename(Chemical_compound = Combination) %>%
  arrange(Sample_ID, Chemical_compound)

res <- ida_valid(mono, combi, listofcombi)
actual <- res[[1]] %>%
  mutate(Combination = paste0(Drug1, " - ", Drug2)) %>%
  relocate(last_col())

# write.csv(actual, "output/oshu_ida_prediction.csv", row.names = F)
```

## outliers/residual

outliers - stand res of more than 3 (with citation) - <https://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/>

R²: 0.5671231 
 <2std.resid >=2std.resid 
         159           11 
         
total: 58/170 had idacomboscore < 0
deviated from 0 (+ve): 7/170 
deviated from 0 (-ve): 4/170 
```{r}
# Compute R² manually
SS_res <- sum((actual$Actual_s_dss - actual$Mean_Combo_s_dss)^2)  # Residual sum of squares
SS_tot <- sum((actual$Actual_s_dss - mean(actual$Actual_s_dss))^2)  # Total sum of squares
R_squared <- 1 - (SS_res / SS_tot)

cat("R²:", R_squared, "\n")

# Sample data of DSS & Compute residuals
residuals <- actual$Actual_s_dss - actual$Mean_Combo_s_dss
actual$`res (mean DSS)` <- residuals
actual$`stand.res (mean DSS)` <- residuals / sd(residuals)
actual$`category (mean DSS)` <- ifelse(abs(actual$`stand.res (mean DSS)`)>=2, ">=2std.resid", "<2std.resid")

# Sample data of IDAComboscore & Compute residuals
residuals <- actual$Actual_IDA_Comboscore - actual$IDA_Comboscore
actual$`res (IDAComboscore)` <- residuals
actual$`stand.res (IDAComboscore)` <- residuals / sd(residuals, na.rm=T)
actual$`category (IDAComboscore)` <- ifelse(abs(actual$`stand.res (IDAComboscore)`)>=2, ">=2std.resid", "<2std.resid")

actual <- actual %>%
  dplyr::select(-`res (mean DSS)`, -`res (IDAComboscore)`) %>%
  mutate(outlier = ifelse(`category (mean DSS)`==">=2std.resid", ifelse(`category (IDAComboscore)`==">=2std.resid", "Both", "mean DSS"), ifelse(`category (IDAComboscore)`==">=2std.resid", "IDAComboscore", NA)))

write.csv(actual, "output/oshu_ida_prediction.csv", row.names = F)
```

# 6. BAML prediction

- more than half of the sample died at the lowest dose in the conc range for elesclomol - hence gonna remove because it is too toxic for the conc range its used in 
```{r}
source("src/function/ida_predict.R")
```

## all samples
- for all: n=580samples, n=165drugs excluding elesclomol
- 10896 predicted combinations
```{r}
data <- read.csv("data/baml/DSS_BAML_postfilt.csv") %>%
  filter(Chemical_compound != "Elesclomol")

output_baml <- ida_predict(Monotherapy_Data = data, output = "output/baml_IDA_prediction.csv", min.sample = 20)
```

## WO overlapping samples with bamldrugcombo
- without samples from drugcombo (n=219) - ie.n=361 overlapping samples
- unique drug (n=142)
- 7618 predicted combinations
```{r}
data <- read.csv("data/baml/DSS_BAML_woOSHU.csv")

output_baml_woOSHU <- ida_predict(Monotherapy_Data = data, output = "output/supp/baml_woOSHU_IDA_prediction.csv", min.sample = 20)
```


# 7. BAML and FP prediction
- overall, FP had overall 515 drugs tested; wo processing - there is 104 common drugs (excluding elesclomol)
  - 330 tested for MCM
- MCM subset had n=74samples, n=79drugs

```{r}
drugdet_fp <- read.csv("data/fpmtb/mono_detail.csv") %>%
  mutate(Approval_status = ifelse(grepl("Inves", Approval_status), "Investigational", Approval_status) )

drugdet <- read.csv("data/baml/mono_detail.csv") %>%
  mutate(Approval_status = ifelse(grepl("Inves", Approval_status), "Investigational", Approval_status) )
```

## all samples
- 27155 combinations 
- Using unique drug: 330; sample: 74
```{r}
data_mcm <- read.csv("data/fpmtb/DSS_FPMTB_MCM_conc_full.csv") # %>%

length(unique(data_mcm$Chemical_compound))

output_fp <- ida_predict(Monotherapy_Data = data_mcm, output = "output/supp/fp_full_IDA_prediction.csv", min.sample = 5) # 5 due to the lowest no of samples tested

```

## for common drugs 
- 80/104 had samples tested, 79/80 drugs have enough sample size
- Using unique drug: 79; sample: 74
```{r}
common_drug <- drugdet_fp$Chemical_compound[drugdet_fp$Chemical_compound %in% drugdet$Chemical_compound] #104

## limiting to common drug: fpmtb
data_mcm <- read.csv("data/fpmtb/DSS_FPMTB_MCM_conc.csv") %>%
  filter(Chemical_compound %in% common_drug) 

output_fp <- ida_predict(Monotherapy_Data = data_mcm, output = "output/supp/fp_commondrug_IDA_prediction.csv", min.sample = 5) # 1933 combinations # 5 due to the lowest no of samples tested

## limiting to common drug: baml
output_baml <- read.csv("output/baml_IDA_prediction.csv") %>%
  filter(Combination %in% output_fp$Combination) %>% 
  group_by(is.na(rank)) %>%
  mutate(rank = as.integer(ifelse(!is.na(rank), rank(-IDA_Comboscore), NA))) 
```

- n=1782 common predicted combination
```{r}
drug_status <- setNames(drugdet$Approval_status, drugdet$Chemical_compound)

output_fp <- read.csv("output/supp/fp_commondrug_IDA_prediction.csv") %>%
  filter(Combination %in% output_baml$Combination) %>% 
  group_by(is.na(rank)) %>%
  mutate(rank = as.integer(ifelse(!is.na(rank), rank(-IDA_Comboscore), NA))) %>%
  rowwise() %>%
  mutate(Approval_status1 = drug_status[Drug1],
         Approval_status2 = drug_status[Drug2],
         approval_status = paste0(pmin(Approval_status1,Approval_status2),"-", pmax(Approval_status1,Approval_status2)) )

combine_df <- left_join(output_baml[, c(14, 1:2, 5:7, 12:13, 15)], output_fp[, c(14, 1:2, 5:7, 12:13, 15, 19)], suffix = c(".baml", ".fp"), by = c("Combination", "Drug1", "Drug2")) %>%
  arrange(rank.baml + rank.fp) %>%
  mutate(# Ensure ranks are numeric (do this first!)
    rank.baml = as.numeric(rank.baml),
    rank.fp = as.numeric(rank.fp),
    category = case_when(
      rank.baml <= 100 & rank.fp <= 100 ~ "Both",
      rank.baml <= 100 & (rank.fp > 100|is.na(rank.fp)) ~ "BeatAML",
      (rank.baml > 100|is.na(rank.baml)) & rank.fp <= 100 ~ "FPMTB-MCM",
      rank.baml > 100 & rank.fp > 100 ~ "none",
      TRUE ~ NA # Default case: if none of the above conditions are met
    )
  ) 

write.csv(combine_df, "output/combine_baml_fp_IDA_prediction.csv", row.names = F)
```

# other synergy models

*** refer to src/analysis/2.synergy_processing.Rmd

# figure 2
- other data from fig 1
```{r}
actual <- read.csv("output/oshu_ida_prediction.csv", check.names = F) 
mono <- read.csv("data/oshu/DSS_mono.csv") 
combi <- read.csv("data/oshu/DSS_combi.csv") 
HSA_eval <- read.csv("output/supp/oshu_HSA_eval.csv", check.names = F)
HSA_df <- HSA_eval %>%
  reframe(median_dss = median(s_dss, na.rm=T), 
          n = n(), 
          median_CI = median(CI, na.rm=T), 
          median_excesshsa = median(Excess_HSA, na.rm=T),
          fill = sum(CI<1 , na.rm = T)*100/n, .by=c(Combination)) 
```

## fig 2b 
```{r}
# variable - combi, mono, actual
p <- mutate(combi[,c(5,8)], Values="Combination") %>%
  dplyr::rename(Chemical_compound = Combination) %>%
  rbind(., mutate(mono[,c(3,6)], Values="Monotherapy")) %>%
  mutate(Values = factor(Values, levels=c("Combination", "Monotherapy")))


p <- p[p$Values=="Monotherapy",c(3,2, 1)] %>%
  reframe(s_dss=mean(s_dss),.by = c(Values, Chemical_compound)) %>% select(-Chemical_compound) %>%
  rbind(., mutate(data.frame(s_dss=actual$Actual_s_dss), Values="Combination \n(Observed)")) %>%
  rbind(., mutate(data.frame(s_dss=actual$Mean_Combo_s_dss), Values="Combination \n(Predicted)")) 

p %>%
  mutate(Values = factor(Values, levels = c("Combination \n(Predicted)", "Combination \n(Observed)", "Monotherapy"))) %>%
ggplot(., aes(x = s_dss, y = Values, fill = Values)) +
  ggridges::stat_density_ridges(
    quantile_lines = TRUE,
    scale = 1,
    rel_min_height = 0.01,
    alpha = 0.7,
    quantiles = 2
  ) +
  scale_fill_manual(
    values = c("#efc000ff","#FC4E07","#0073c2ff"),
    guide = "none"
  ) +
  stat_compare_means(
    comparisons = list(
      c("Monotherapy", "Combination \n(Observed)"),
      c("Combination \n(Predicted)", "Combination \n(Observed)"),
      c("Monotherapy", "Combination \n(Predicted)")
    ),
    label = "p.signif",  
    method = "wilcox.test",
    label.x = c(0.82, 0.87, 0.94)  
  ) +
  labs(x="Mean DSS", y="") +
  theme_classic2(base_size = 14) +
  theme(aspect.ratio = 0.5,
        panel.grid.major.y = element_line(color = "#A8BAC4", linewidth = 0.3) ) 

ggsave("plot/fig3/main_3b.png", width = 5, height = 3, units = "in", dpi=600)
```

## fig 2c
```{r}
# variable - actual, HSA_eval
df <- HSA_eval %>% 
  group_by(Combination) %>%
  cor_test(vars = "HSA_ref", vars2 = "s_dss", method = "spearman") %>%
  left_join(actual) %>%
  mutate(venaza = ifelse(Combination == "Azacitidine - Venetoclax", Combination, NA)) 

PvA_corr.plot(df, xlim=c(0,1), ylim=c(0,1), color = "cor", size = "n", save="plot/fig3/main_3c.png", width = 26, height = 24, dpi = 600)
```

## supp fig 3a
```{r eval=FALSE}
# variable - actual

## standardised residual vs fitted
p1 <- ggplot(actual, aes(x = Mean_Combo_s_dss, y = `stand.res (mean DSS)`, color=`category (mean DSS)`)) +
  geom_point() +  # Scatterplot of residuals
  scale_color_manual(
    values =  c(">=2std.resid" = "red","<2std.resid" = "black"), 
    labels = c(expression("|r"[i] * "|" < 2), expression("|r"[i] * "|" >= 2)),
    name = "Stand.Resid") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  # Reference line at 0
  geom_smooth(method = "loess", color = "blue", se = FALSE) +  # Fitted trend line
  labs(x = "Fitted (Predicted) Values",
       y = "Standardized Residuals") +
  xlim(c(0,1)) +
  theme_classic2() + theme(aspect.ratio = 1, legend.position = "bottom")

```


```{r}
# variable - actual

plot <- actual %>%
  mutate(label = ifelse(abs(`stand.res (mean DSS)`)>=2, Combination, NA)) %>%
  dplyr::rename(colour = `category (mean DSS)`) 

p2 <- ggplot(plot, aes(x=Actual_s_dss, y = Mean_Combo_s_dss, color=colour, label=label)) +
  geom_point() + 
  geom_smooth(method = "lm", color="blue", se=FALSE) +
  scale_color_manual(
    values =  c(">=2std.resid" = "red","<2std.resid" = "black"), 
    labels = c(expression("|r"[i] * "|" < 2), expression("|r"[i] * "|" >= 2)),
    name = "Stand.Resid") +
  geom_text_repel(color = "grey29",box.padding = unit(1.2, "lines"), size=3,
                  point.padding = unit(0.5, "lines"),
                  segment.color = "black", min.segment.length = 0,
                  force = 1, max.overlaps = 20, seed = 123) +
  stat_cor(data=plot[plot$colour == ">=2std.resid",],
           aes(x=Actual_s_dss, y=Mean_Combo_s_dss), method = "spearman", 
           label.x.npc = "left", label.y = 0.9, color="red", 
           inherit.aes = FALSE, show.legend=FALSE)  +
  stat_cor(data=plot[plot$colour == "<2std.resid",], 
           aes(x=Actual_s_dss, y=Mean_Combo_s_dss), method = "spearman",
           label.x.npc = "left", label.y = 0.95, color="black", 
           inherit.aes = FALSE, show.legend=FALSE) + 
  xlim(c(0,1)) + ylim(c(0,1)) +
  theme_classic2() +
  labs(x = "Observed Mean Combo DSS", y = "Predicted Mean Combo DSS") + theme(aspect.ratio = 1, legend.position = "bottom")

```

```{r}
p1 / p2 + plot_layout(guides = "collect")  &
  theme(legend.position='bottom')

ggsave("plot/suppfig3/supp_3a.png", units = "cm", width = 13, height = 26)
```

## supp fig 3b 
- 112 combinations with values observed idacomboscore
```{r }
# variable - actual
plot <- actual %>%
  mutate( label3 = ifelse(!is.na(outlier), Combination, NA) ) %>%
  filter(!(Actual_HR_vs_Drug1 >= 1 | Actual_HR_vs_Drug2 >= 1))

# 112 combinations
p1 <- ggplot(plot, aes(x = Actual_IDA_Comboscore, y = IDA_Comboscore, color=`category (mean DSS)`, label = label3)) +
  geom_point(size = 1) +  # Scatterplot of residuals
  scale_color_manual(
    values =  c(">=2std.resid" = "red","<2std.resid" = "black"), 
    labels = c(expression("|r"[i] * "|" < 2), expression("|r"[i] * "|" >= 2)),
    name = "Stand.Resid") +
  geom_abline(intercept = 0) +
  geom_text_repel(segment.color = "black", min.segment.length = 0.1,
      max.overlaps = 8, seed = 123, size = 3, nudge_y = 0.03, force = 3) +
  xlim(c(0,0.3)) + ylim(c(0,0.3)) +
  labs(x = "Observed IDA Comboscore",
       y = "Predicted IDA Comboscore") +
  theme_classic() + theme(aspect.ratio = 1, legend.position = "bottom")

ggExtra::ggMarginal(p1, type = "densigram", size=7)
ggsave("plot/suppfig3/supp_3b.png", width = 6, height = 5, units = "in", dpi = 600)
```

## supp fig 3c: comparison with other synergy
```{r}
load("output/supp/Bliss_ref_list.RData")
load("output/supp/Excess_bliss_list.RData")

bliss_df <- data.frame(
  Sample_ID = Bliss_ref_list$Sample_ID, 
  Combination = Bliss_ref_list$Combination, 
  CI = Bliss_ref_list$DSS/Bliss_ref_list$s_dss, 
  excess_bliss = Bliss_ref_list$s_dss - Bliss_ref_list$DSS, 
  s_dss = Excess_bliss_list$s_dss 
  ) %>%
  group_by(Combination) %>%
  reframe(median_dss = median(s_dss, na.rm=T), n=n(), 
          median_CI = median(CI, na.rm=T),
          median_excessbliss = median(excess_bliss, na.rm=T)) 
```

```{r}
load("output/supp/Loewe_ref_list.RData")
load("output/supp/Excess_loewe_list.RData")

loewe_df <- data.frame(
  Sample_ID = Loewe_ref_list$Sample_ID, 
  Combination = Loewe_ref_list$Combination, 
  CI = Loewe_ref_list$DSS/Loewe_ref_list$s_dss, 
  excess_loewe = Loewe_ref_list$s_dss - Loewe_ref_list$DSS, 
  s_dss = Excess_loewe_list$s_dss 
  ) %>%
  group_by(Combination) %>%
  reframe(median_dss=median(s_dss, na.rm=T), n=n(), 
          median_CI = median(CI, na.rm=T),
          median_excessloewe=median(excess_loewe, na.rm=T)) 
```

```{r}
all_df <- full_join(HSA_df[1:5], bliss_df, by=c("Combination", "median_dss", "n")) %>%
  inner_join(loewe_df, by=c("Combination", "median_dss", "n")) %>%
  inner_join(df[,c("Combination", "cor", "stand.res (mean DSS)")] ) %>%
  dplyr::rename(stand.res = `stand.res (mean DSS)`) %>%
  arrange(desc(stand.res)) %>%
  mutate(Label = ifelse(abs(stand.res) >= 2, Combination, NA))

colnames(all_df)[c(4,6,8)] <- c("median_hsa_CI", "median_bliss_CI", "median_loewe_CI")
```

```{r}
wrapper_plot <- function(yvar, ylab_text) {
  PvA_corr.plot(
    all_df,
    x = "median_excesshsa", y = yvar,
    color = "stand.res", colorname = "Stand.Resid", size = 3,
    xlab = "Median Excess (HSA)", ylab = ylab_text,
    xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5),
    label1 = NA, label2 = NA,
    palette = c("#0571B0", "grey80", "#CA0020")
  ) +
    geom_abline(linetype = "dashed") +
    geom_text_repel(
      aes(label = Label),
      size = 4, direction = "both",
      min.segment.length = 0.5, box.padding = 0.25,
      point.padding = unit(0.2, "lines"), max.overlaps = 10,
      force = 1, seed = 123
    ) +
    theme(aspect.ratio = 1, legend.position = c(0.9, 0.2), legend.background = element_rect(fill = "transparent", colour = NA))
}

# Create the two plots
p1 <- wrapper_plot("median_excessloewe", "Median Excess (Loewe)")
p2 <- wrapper_plot("median_excessbliss", "Median Excess (Bliss)")
p1 / p2

ggsave("plot/suppfig3/supp_3c.png", dpi = 600, units = "cm", width = 15, height = 30)
```


# figure 3

```{r}
library(VennDiagram)
drugdet <- read.csv("data/oshu/mono_detail.csv")
listofdrug <- drugdet[!duplicated(drugdet$Drug),] %>%
  filter(! Drug %in% c("Arsenic Trioxide", "Quizartinib - NVP-TAE684"))
actual <- read.csv("output/oshu_ida_prediction.csv", check.names = F) 

baml_drugdet <- read.csv("data/baml/mono_detail.csv") 
baml_mono <- read.csv("data/baml/DSS_BAML_postfilt.csv") %>%
  filter(Chemical_compound != "Elesclomol") # wasnt used

fp_mono_all <- read.csv("data/fpmtb/DSS_FPMTB_MCM_conc_full.csv")
fp_drugdet <- read.csv("data/fpmtb/mono_detail.csv") %>%
  filter(Chemical_compound %in% unique(fp_mono_all$Chemical_compound))
fp_mono <- read.csv("data/fpmtb/DSS_FPMTB_MCM_conc.csv")

baml_woOSHU_output <- read.csv("output/supp/baml_woOSHU_IDA_prediction.csv")
baml_output <- read.csv("output/baml_IDA_prediction.csv")
bamlfp_output <- read.csv("output/combine_baml_fp_IDA_prediction.csv")
fp_all_output <- read.csv("output/supp/fp_full_IDA_prediction.csv")
```

## fig 3f: venn
```{r message=FALSE}
# Chart
venn.diagram(
        x = list(fp_drugdet[["Chemical_compound"]], baml_drugdet[["Chemical_compound"]]),
        category.names = c("FPMTB-MCM" , "BeatAML"),
        filename = 'plot/fig3/main_3f_venn.svg',
        disable.logging = T,
        
        # Output features
        imagetype="svg" ,
        units = "in",
        height = 2 , 
        width = 2 , 
        resolution = 600,
        # compression = "lzw",
        
        # Circles
        lwd = 1,
        # lty = 'blank',
        col = "black",
        fill = c("#E41A1C","#377EB8"),
        alpha = 0.5,
        
        # Numbers
        cex = .6,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-22, 8),
        cat.dist = c(0.055, 0.055),
        cat.fontfamily = "sans"
)
```

## fig 3d: top drug parter based on median predicted dss of all predicted combinations they are part of
```{r}
# variable - baml_output
# Step 1: Create symmetric long-form data
long_df <- baml_output %>%
  select(Drug1, Drug2, Score = Mean_Combo_s_dss) %>%
  bind_rows(., rename(., Drug1 = Drug2, Drug2 = Drug1))  # Ensure symmetry

# Step 2: Remove self-pairs
long_df <- long_df %>% filter(Drug1 != Drug2) %>% rename(Drug := Drug1)

# Compute row means
row_medians <- long_df %>%
  group_by(Drug) %>%
  summarise(median_score = median(Score)) %>%
  arrange(desc(median_score))

long_df %>%
  filter(Drug %in% row_medians$Drug[1:20]) %>%
  mutate(Drug=factor(Drug, levels=rev(row_medians$Drug[1:20]))) %>%
  ggplot(., aes(x=Drug, y=Score, fill=Score)) +
  ggbeeswarm::geom_beeswarm(cex = 2.5, corral = "wrap", corral.width = 0.7, shape=21, alpha=0.7) +
  stat_summary(fun = median, geom = "crossbar", width = 0.9,
               color = "black", fatten = 1, linewidth = 0.8) +
  scale_fill_distiller(palette = "RdYlBu", name = "DSS") +
  labs(x="Top Drug Partners", y="Predicted Combination DSS", subtitle = "Beat AML") +
  coord_flip() +
  theme_classic2(base_size = 12) +
  theme(aspect.ratio = 0.7, 
        axis.text = element_text(colour = "black", size = 14),
        panel.grid.major.x = element_line(color = "#A8BAC4"))

ggsave("plot/fig3/main_3d.png", width = 8, height = 6, units = "in", dpi = 600)
```

## fig 3e: top drug partner based on frequency of occurrence in top 100 combinations
```{r}
top_100 <- baml_output %>%
  slice_min(rank, n = 100)

tables <- annotate_drug_classes(top_100, baml_drugdet[,c(2,5:8)], "Chemical_compound")

combo_metadata <- tables[[1]]
combo_class_count <- tables[[2]]
combo_class_labels <- tables[[3]]

drug_counts <- count_unique_drugs(combo_metadata)

plot_top_drugs(drug_counts, top_n = 20, subtitle = "Beat AML") 

ggsave(paste0("plot/fig3/main_3e_topdrugpartner.png"), units="in", width=8, height=5, dpi=400)
```

## fig 3f: top 10 combinations with approved-approved drug partners

```{r}
# Helper function for monotherapy DSS per sample
process_dataset <- function(data_df, selected, dataset_label) {
  rbindlist(
    lapply(seq_len(nrow(selected)), function(i) {
      drugs <- as.character(selected[i, 2:3])
      combo <- selected$Combination[i]

      data_df %>%
        filter(Chemical_compound %in% drugs) %>%
        reshape2::dcast(Sample_ID ~ Chemical_compound, fun.aggregate = mean, 
              value.var = "s_dss", fill = NA_real_) %>%
        filter(complete.cases(.)) %>%
        setnames(old = drugs, new = c("Drug1_dss", "Drug2_dss")) %>%
        mutate(
          best_drug = case_when(
            Drug1_dss == Drug2_dss & Drug1_dss == 0 ~ "0",
            Drug1_dss == Drug2_dss ~ "Both",
            Drug1_dss < Drug2_dss ~ "Drug2",
            TRUE ~ "Drug1"
          ),
          HSA_ref = pmax(Drug1_dss, Drug2_dss),
          Drug1 = drugs[1],
          Drug2 = drugs[2],
          Combination = combo,
          dataset = dataset_label
        )
    })
  )
}

selected <- bamlfp_output %>% filter(category == "Both", approval_status == "Approved-Approved") %>%
  arrange(Mean_Combo_s_dss.baml)


data_baml <- read.csv("data/baml/DSS_BAML_postfilt.csv") %>%
  filter(Chemical_compound %in% unique(fp_mono$Chemical_compound))

# Process both datasets
baml <- process_dataset(data_baml, selected, "BeatAML")
fp   <- process_dataset(fp_mono, selected, "FPMTB-MCM")

# Compute best drug proportions
df_summary <- rbind(baml, fp) %>%
  group_by(Combination, Drug1, Drug2, dataset) %>%
  summarise(
    n = n(),
    Drug1_percent = mean(best_drug == "Drug1") * 100,
    Drug2_percent = mean(best_drug == "Drug2") * 100,
    .groups = "drop"
  )

# Plot
rbind(baml, fp) %>%
  left_join(df_summary, by = c("Combination", "Drug1", "Drug2", "dataset")) %>%
  mutate(Combination = factor(Combination, levels = selected$Combination)) %>%
  ggplot(aes(y = Combination, x = HSA_ref, fill = Drug1_percent)) +
  ggridges::geom_density_ridges(quantile_lines = TRUE, scale = 1.5, rel_min_height = 0.01, alpha = 0.7, quantiles = 2) +
  scale_fill_gradient2(
    midpoint = 50,
    high = "#FC8D59", mid = "#F7F7F7", low = "#7FBF7B",
    name = "%Sample\npredicted\nas Drug1"
  ) +
  labs(x = "Predicted DSS", y = "") +
  facet_wrap(~dataset, scales = "free_x") +
  ggridges::theme_ridges() +
  theme(axis.title.x = element_text(hjust = 0.5), aspect.ratio = 1.4, legend.title = element_text(size = 10))


ggsave(paste0("plot/fig3/main_3f.png"), units="in", width=8, height=5, dpi=400)
```
## suppfig 4a (left)

```{r}
venn.diagram(
        x = list(baml_drugdet[["Chemical_compound"]], listofdrug[["Drug"]]),
        category.names = c("BeatAML" , "BeatAML-\nDrugCombo"),
        filename = 'plot/suppfig4/suppfig_4a_venn.png',
        disable.logging = T,
        
        # Output features
        imagetype="png" ,
        height = 1000 , 
        width = 1000 , 
        resolution = 600,
        # compression = "lzw",
        
        # Circles
        lwd = 1,
        # lty = 'blank',
        col = "black",
        fill = c("#377EB8", "#4DAF4A"),
        alpha = 0.5,
        
        # Numbers
        cex = .6,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-22, 8),
        cat.dist = c(0.055, 0.055),
        cat.fontfamily = "sans"
)

```


## suppfig 4a (right)
```{r}
actual %>%
  dplyr::select(Combination, Actual_s_dss) %>%
  inner_join(., baml_woOSHU_output, by="Combination", keep = T) %>%
  ggscatter(., x="Actual_s_dss", y="Mean_Combo_s_dss",
        add = "reg.line", conf.int = F,
        size=1.5, 
        cor.coef = TRUE, cor.method = "spearman",
        add.params = list(color = "blue", fill = "lightgray"),
        xlab = "Observed Mean Combo DSS\n(BeatAML-DrugCombo)", 
        ylab = "Predicted Mean Combo DSS\n(BeatAML)",
        title = paste0("n=",nrow(.))
        ) +
  theme_classic2() + theme(aspect.ratio = 1)
  
ggsave("plot/suppfig4/suppfig_4a_right.png", width = 9, height = 9, units = "cm", dpi=600)
```

## suppfig 4b
```{r}
# variable - baml_drugdet, baml_mono
drugclassdet <- setNames(baml_drugdet$Drug_Class, baml_drugdet$Chemical_compound)

plot <- baml_mono %>%
  group_by(Chemical_compound) %>%
  reframe(pt25 = quantile(s_dss,probs=.25),
            median = median(s_dss),
            pt75 = quantile(s_dss, probs=.75),
            n=n()) %>%
  mutate(label = paste0(Chemical_compound, " (n=", n, ")"),
         Drug_Class = drugclassdet[Chemical_compound],
         drug1_colour = drugclass[Drug_Class] ,
         label.format = paste0("<span style='color:", drug1_colour,";'>", Chemical_compound,"</span> (n=",n,")</span>")) %>%
  slice_max(order_by = median, n = 20)


library(ggtext)
ggplot(plot, aes(x = reorder(label, median, decreasing = F), y = median, ymin = pt25, ymax = pt75, color = Drug_Class) ) + 
  geom_errorbar() +
  geom_point() +
  scale_colour_manual(values = drugclass) +
  scale_x_discrete(labels = setNames(plot$label.format, plot$label))+
  labs(x="Top Monotherapies", y="DSS", subtitle = "Beat AML") +
  theme_pubr() +
  coord_flip() +
  theme(axis.text.y = ggtext::element_markdown(angle = 0, hjust = 1, vjust = 0.5),
        legend.text = element_text(size=8),
        aspect.ratio = 0.7, legend.position = c(0.95, 0.2),
        panel.grid.major.x = element_line(color = "#A8BAC4"),
        legend.key.height = unit(0.1, "cm")
        # legend.background = element_rect(fill = NA)
        ) 

ggsave("plot/suppfig4/supp_4b.png", width = 9, height = 4, units = "in", dpi=600)
```

## suppfig 4d
```{r}
temp <- data.frame(Group="Monotherapy", Value=baml_mono$s_dss, Chemical_compound=baml_mono$Chemical_compound) %>%
  summarise(Value = mean(Value), .by = c(Group, Chemical_compound)) %>%
  select(-Chemical_compound)

temp <- data.frame(Group="Predicted\nCombination", Value=baml_output$Mean_Combo_s_dss) %>%
  rbind(temp) %>%
  mutate(Group=factor(Group, levels=c("Predicted\nCombination", "Monotherapy")))

pval <- compare_means(Value ~ Group, data = temp, method = "wilcox.test") #<2e-16

p1 <- ggplot(temp , aes(y=Group,x=Value, fill=Group)) +
  scale_fill_manual(values=c("#efc000ff", "#0073c2ff"), guide="none") +
  ggridges::geom_density_ridges(quantile_lines=TRUE, scale = 1, rel_min_height = 0.01, alpha=0.7, quantiles=2)  + 
  labs(x="Mean DSS", y="", title = "Beat AML") +
  theme_classic2(base_size = 14) +
  theme(aspect.ratio = 0.4, 
        panel.grid.major.y = element_line(color = "#A8BAC4", size = 0.3) ) + 
  annotate("text", x = max(temp$Value) * 0.95, y = 2.8, label = "p < 2.2e-16", hjust = 1, size = 4)
```

```{r}
temp <- data.frame(Group="Monotherapy", Value=fp_mono_all$s_dss, Chemical_compound=fp_mono_all$Chemical_compound) %>%
  summarise(Value = mean(Value), .by = c(Group, Chemical_compound)) %>%
  select(-Chemical_compound)

temp <- data.frame(Group="Predicted\nCombination",
                   Value=fp_all_output$Mean_Combo_s_dss
                   # Value=bamlfp_output$Mean_Combo_s_dss.fp
                   ) %>%
  rbind(temp) %>%
  mutate(Group=factor(Group, levels=c("Predicted\nCombination", "Monotherapy")))

pval <- compare_means(Value ~ Group, data = temp, method = "wilcox.test") #2.5e-11

p2 <- ggplot(temp , aes(y=Group,x=Value, fill=Group)) +
  scale_fill_manual(values=c("#efc000ff", "#0073c2ff"), guide="none") +
  ggridges::geom_density_ridges(quantile_lines=TRUE, scale = 1, rel_min_height = 0.01, alpha=0.7, quantiles=2)  + 
  labs(x="Mean DSS", y="", title = "FPMTB-MCM") +
  theme_classic2(base_size = 14) +
  theme(aspect.ratio = 0.4,
        panel.grid.major.y = element_line(color = "#A8BAC4", size = 0.3) ) + 
  annotate("text", x = max(temp$Value) * 0.95, y = 2.8, label = "p < 2.5e-11", hjust = 1, size = 4)
```

```{r}
p1 / p2
ggsave("plot/suppfig4/supp_4d.png", width = 5, height = 5, units = "in", dpi = 600)
```

## suppfig 4e
```{r}
bamlfp_output %>%
  ggscatter(., x="Mean_Combo_s_dss.baml",
          y="Mean_Combo_s_dss.fp",
          add = "reg.line", 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "black"),
          color = "#8DA0CB", alpha = 0.7, size = .8,
          xlab = "Predicted Mean Combo DSS\n(BeatAML)", 
          ylab = "Predicted Mean Combo DSS\n(FPMTB-MCM)",
          title = paste0("n=",nrow(.))
          ) +
  theme_classic2() + theme(aspect.ratio = 1)
  
ggsave("plot/suppfig4/supp_4e.png", width = 9, height = 9, units = "cm", dpi=600)
```

## suppfig 4f
```{r}
summary_df <- bamlfp_output %>%
  filter(!is.na(category), category != "none") %>%
  select(category, approval_status) %>%
  mutate(approval_status = gsub("-", "-\n", approval_status)) %>%
  group_by(category, approval_status) %>%
  count(approval_status) %>%
  arrange(desc(n)) %>%
  mutate(category = factor(category, levels = c("FPMTB-MCM", "Both", "BeatAML")) ) %>%
  arrange(desc(approval_status))

summary_df$approval_status <- factor(summary_df$approval_status, levels =  unique(summary_df$approval_status))

p <- ggplot(summary_df, aes(y = approval_status, x = n, fill = category)) +
  geom_col( width = 0.7) +
  geom_text(aes(label = n), 
            position = position_stack(vjust = 0.5), 
            size = 3.5) +
  scale_fill_manual(values = c("#e06666","#8e7cc3","#6d9eeb"), name = NULL) +
  scale_x_continuous(expand = c(0,0.1), limits = c(0,90)) +
  labs( y = "Approval status", x = "No. of predicted combinations" ) +  
  theme_classic(base_size = 14) +
  theme(
    aspect.ratio = 0.6,
    legend.position = "top",
    axis.text = element_text(color = "black"),
    panel.grid.major.x = element_line(color = "#A8BAC4", size = 0.3),
    axis.ticks.length = unit(0, "mm"),
    plot.title = element_text(face = "bold")
  ) 

p +
  geom_text(data = summary_df %>% 
              group_by(approval_status) %>% 
              summarise(n = sum(n)), 
            aes(y = approval_status, x = n, label = n),
            hjust = -0.8,
            size = 3.5, inherit.aes = F)

ggsave("plot/suppfig4/supp_4f.png", width = 6, height = 4, units = "in", dpi = 600)
```

# sessioninfo
```{r}
sessionInfo()
```

