---
title: "Synergy Calculation"
description: "not processed in this github"
date: "23 February | 26 March | 21 may"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---

- purpose: to see if bad predictions can be explained by drug interaction
- code for data for Bliss/Loewe are from synergyfinder

code:
<https://github.com/shuyuzheng/synergyfinder>

vignette:
<https://www.bioconductor.org/packages/release/bioc/vignettes/synergyfinder/inst/doc/User_tutorual_of_the_SynergyFinder_plus.html>

tutorial: <https://europepmc.org/article/PMC/6383747>

# loading library
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
source("src/function/supp/loewe_fn.R")
source("src/function/supp/curvefit.R")
```

# loading data
```{r}
mono <- read.csv("data/oshu/DSS_mono.csv")
combi <- read.csv("data/oshu/DSS_combi.csv")

listofcombi <- unique(combi[,3:5])

combidet <- read.csv("data/oshu/combi_detail.csv")[3:4]
colnames(combidet) <- c("name", "Chemical_compound")
drugdet <- read.csv("data/oshu/mono_detail.csv")[1:2]
colnames(drugdet) <- c("name", "Chemical_compound")

all_drugdet <- rbind(combidet, drugdet)

sampleinfo <- read.csv("data/oshu/clinical_summary.csv")[,c(3,4,2)] %>%
  dplyr::rename(lab_id = labId, patient=patientId)

raw_data <- read.csv("data/oshu/raw/viability_oshu.csv")[,c(1,2,6,7,10,11,14)] %>%
  filter(drug != "Control",
         drug != "Quizartinib - NVP-TAE684 - Arsenic Trioxide") %>%
  mutate(percent_viability = normalized_probit_result/100,
         percent_inhibition = 1 - normalized_probit_result/100) %>%
  left_join(., all_drugdet, by=c("drug" = "name"), multiple = "first") %>%
  inner_join(., sampleinfo) 

```

# combination index (highest single agent)
- function:
  CI = max(E_a, E_b) / E_ab 
    CI < 1 = combination more effective 
    CI = 1 = equally effective 
    CI > 1 = monotherapy is better
  Excess HSA = Observed - Expected
- using DSS as efficacy
- summarized using median over samples

# HSA/IDA's calculation
```{r}
HSA_eval <- rbindlist(
  lapply(setNames(1:170, listofcombi$Combination), function(i){
    
    # combination DSS per sample
    data <- combi[combi$Combination==listofcombi[i, 3],] %>%
      dplyr::select(Sample_ID, Combination, s_dss)
    
    # monotherapy DSS per sample
    data <- mono[mono$Chemical_compound %in% listofcombi[i, 1:2],] %>%
      reshape2::dcast(Sample_ID ~ Chemical_compound, fun.aggregate = mean, value.var = "s_dss", fill = NA_real_) %>%
      inner_join(., data, by="Sample_ID")
    
    data <- data[complete.cases(data),]
    colnames(data)[2:3] <- c("Drug1_dss", "Drug2_dss")
    
    data$best_mono <- apply(data[,2:3], 1, function(x){ifelse(x[1]==x[2], NA, listofcombi[i, which.max(x)])})
    data$HSA_ref = pmax(data$Drug1_dss, data$Drug2_dss)
    data$Excess_HSA = data$s_dss - data$HSA_ref
    data$CI <- data$HSA_ref/data$s_dss
    data$CI[is.nan(data$CI)] = NA # Inf when combi=0; NaN when all is 0
    data
  })
)

# save(HSA_eval, file="output/synergy/HSA_eval.RData")
```

```{r}
##### HSA concentration method (not evaluated) #####
## HSA calculated at each conc point before recalculating DSS
# HSA_ref_list = list()
# HSA_CI_list = list()
# Excess_hsa_list = list()
# pb <- txtProgressBar(min = 0, max = 170, style = 3, width = 50, char = "=")
# for (i in 1:170){
#   combi_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 3],] %>%
#     reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
#   
#   drug1_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 1],] %>%
#     reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
#   
#   drug2_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 2],] %>%
#     reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
#   
#   common <- Reduce(intersect, list(rownames(combi_), rownames(drug1_), rownames(drug2_)))
#   
#   HSA_ref = pmax(drug1_[common,],drug2_[common,])
#   
#   # DSS for HSA ref
#   DSS <- rbindlist(lapply(common, function(x){
#     suppressWarnings(CALC_DSS_single(HSA_ref[x,]*100, as.numeric(colnames(HSA_ref)), listofcombi[i, 3]))
#   }))
#   colnames(HSA_ref) = paste0("Conc_", 1:7)
#   
#   CI = HSA_ref / combi_[common,]
#   colnames(CI) = paste0("Conc_", 1:7)
#   
#   Excess_hsa = combi_[common,] - HSA_ref
#   colnames(Excess_hsa) = paste0("Conc_", 1:7)
#   
#   HSA_ref = as.data.frame(HSA_ref) 
#   HSA_ref$Sample_ID = rownames(HSA_ref)
#   HSA_ref <- cbind(HSA_ref, DSS)
#   HSA_ref_list [[listofcombi[i, 3]]] = HSA_ref[,c(8,1:7,9:11)]
#   
#   CI = as.data.frame(CI) 
#   CI$Sample_ID = rownames(CI)
#   HSA_CI_list [[listofcombi[i, 3]]] = CI[,c(8,1:7)]
#   
#   Excess_hsa = as.data.frame(Excess_hsa) 
#   Excess_hsa$Sample_ID = rownames(Excess_hsa)
#   Excess_hsa_list [[listofcombi[i, 3]]] = Excess_hsa[,c(8,1:7)]
#   
#   setTxtProgressBar(pb, i)
# }
# close(pb)
# 
# HSA_ref_list = rbindlist(HSA_ref_list, idcol=TRUE) %>%
#   dplyr::rename(Combination = .id) %>%
#   inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
# 
# HSA_CI_list = rbindlist(HSA_CI_list, idcol=TRUE) %>%
#   dplyr::rename(Combination = .id) %>%
#   inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
# HSA_CI_list$CI = apply(HSA_CI_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))
# 
# Excess_hsa_list = rbindlist(Excess_hsa_list, idcol=TRUE) %>%
#   dplyr::rename(Combination = .id) %>%
#   inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
# Excess_hsa_list$excess_hsa = apply(Excess_hsa_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))
# 
# 
# save(HSA_ref_list, file="output/synergy/HSA_ref_list.RData")
# save(HSA_CI_list, file="output/synergy/HSA_CI_list.RData")
# save(Excess_hsa_list, file="output/synergy/Excess_hsa_list.RData")
```


# Bliss's calculation
- function: 
  CI = (E_a + E_b - E_a * E_b) / E_ab
  CI < 1 == synergy
  Excess bliss = Observed - Expected (ideal is expected > observed)
- E refers to *percent inhibition* (which normalised based on neg and pos control)
- higher inhibition == better; so CI < 1 is synergistic

c('drug', 'Concentration','Patient.num' ,'inhibition_percent', 'ID')
```{r}
Bliss_ref_list = list()
Bliss_CI_list = list()
Excess_bliss_list = list()

pb <- txtProgressBar(min = 0, max = 170, style = 3, width = 50, char = "=")
for (i in 1:170){
  combi_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 3],] %>%
    reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
  
  drug1_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 1],] %>%
    reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
  
  drug2_ = raw_data[raw_data$Chemical_compound %in% listofcombi[i, 2],] %>%
    reshape2::acast(Sample_ID ~ concentration, fun.aggregate = mean, value.var = "percent_inhibition", fill = NA_real_) 
  
  common <- Reduce(intersect, list(rownames(combi_), rownames(drug1_), rownames(drug2_)))
  
  Bliss_ref = drug1_[common,] + drug2_[common,] - (drug1_[common,] * drug2_[common,])
  
  # DSS for bliss ref
  DSS <- rbindlist(lapply(common, function(x){
    suppressWarnings(CALC_DSS_single(Bliss_ref[x,]*100, as.numeric(colnames(Bliss_ref)), listofcombi[i, 3]))
  }))
  colnames(Bliss_ref) = paste0("Conc_", 1:7)
  
  CI = Bliss_ref / combi_[common,]
  colnames(CI) = paste0("Conc_", 1:7)
  
  Excess_bliss = combi_[common,] - Bliss_ref
  colnames(Excess_bliss) = paste0("Conc_", 1:7)
  
  Bliss_ref = as.data.frame(Bliss_ref) 
  Bliss_ref$Sample_ID = rownames(Bliss_ref)
  Bliss_ref <- cbind(Bliss_ref, DSS)
  Bliss_ref_list [[listofcombi[i, 3]]] = Bliss_ref[,c(8,1:7,9:11)]
  
  CI = as.data.frame(CI) 
  CI$Sample_ID = rownames(CI)
  Bliss_CI_list [[listofcombi[i, 3]]] = CI[,c(8,1:7)]
  
  Excess_bliss = as.data.frame(Excess_bliss) 
  Excess_bliss$Sample_ID = rownames(Excess_bliss)
  Excess_bliss_list [[listofcombi[i, 3]]] = Excess_bliss[,c(8,1:7)]
  
  setTxtProgressBar(pb, i)
}
close(pb)


Bliss_ref_list = rbindlist(Bliss_ref_list, idcol=TRUE) %>%
  dplyr::rename(Combination = .id) %>%
  inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))

Bliss_CI_list = rbindlist(Bliss_CI_list, idcol=TRUE) %>%
  dplyr::rename(Combination = .id) %>%
  inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
Bliss_CI_list$CI = apply(Bliss_CI_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))

Excess_bliss_list = rbindlist(Excess_bliss_list, idcol=TRUE) %>%
  dplyr::rename(Combination = .id) %>%
  inner_join(., HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
Excess_bliss_list$excess_bliss = apply(Excess_bliss_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))


save(Bliss_ref_list, file="output/supp/Bliss_ref_list.RData")
save(Bliss_CI_list, file="output/supp/Bliss_CI_list.RData")
save(Excess_bliss_list, file="output/supp/Excess_bliss_list.RData")
```


# Loewe's calculation
- function: 
  CI = dose_a/ dose_a(that can give E_ab) + dose_b/ dose_b(that can give E_ab)
  CI < 1 == synergy
- uses *percent inhibition* for E - higher the better
- excess loewe = observed - expected (ideal is observed > expected)
- requires identification of the dose-response curve function
- using functions from synergyfinder instead (modified)
```{r}
# raw_data 
xpr_tbl <- raw_data %>%
  mutate(dose = concentration * 1000,
         response = 100 - normalized_probit_result) %>% # percent inhibition
  dplyr::select(Sample_ID, Chemical_compound, dose, response) %>%
  group_by(Sample_ID, Chemical_compound, dose) %>%
  summarise(response = mean(response)) # averaging replicates 

options(warn = -1)
Loewe_ref_list = list()
Loewe_CI_list = list()
Excess_loewe_list = list()
# Initializes the progress bar
pb <- txtProgressBar(min = 0, max = 170, style = 3, width = 50, char = "=")
for (i in 1:170){
  drug1_ = xpr_tbl[xpr_tbl$Chemical_compound %in% listofcombi[i, 1],] 
  drug2_ = xpr_tbl[xpr_tbl$Chemical_compound %in% listofcombi[i, 2],]
  combi_ = xpr_tbl[xpr_tbl$Chemical_compound %in% listofcombi[i, 3],] 
  
  common <- Reduce(intersect, list(unique(combi_$Sample_ID), unique(drug1_$Sample_ID), unique(drug2_$Sample_ID)))
  
  combi_ <- combi_[combi_$Sample_ID %in% common,]
  drug1_ <- drug1_[drug1_$Sample_ID %in% common,]
  drug2_ <- drug2_[drug2_$Sample_ID %in% common,]
  
  Loewe_ref = list()
  Loewe_CI = list()
  Excess_loewe = list()
  for (j in common){
    drug1 = drug1_[drug1_$Sample_ID == j, c("dose", "response")]
    drug2 = drug2_[drug2_$Sample_ID == j, c("dose", "response")]
    single_drug_data <- list( drug1, drug2 )
    
    response = data.frame(conc1 = drug1$dose,
                          conc2 = drug2$dose,
                          response =  combi_$response[combi_$Sample_ID == j])
    
    res <- Loewe_2(single_drug_data, response)
    
    DSS <- suppressWarnings(CALC_DSS_single(res$Loewe_ref, pmax(response[,1], response[,2]), listofcombi[i, 3]))
    
    Loewe_ref[[j]] = cbind(as.data.frame(t(res$Loewe_ref)), DSS)
    Loewe_CI[[j]] = as.data.frame(t(res$CI)) # returns Inf when E_ab = 0
    Excess_loewe[[j]] = as.data.frame(t(res$Excess_loewe))
  }
  
  Loewe_ref <- rbindlist(Loewe_ref, idcol=TRUE)
  Loewe_CI <- rbindlist(Loewe_CI, idcol=TRUE)
  Excess_loewe <- rbindlist(Excess_loewe, idcol=TRUE)
  
  Loewe_ref_list [[listofcombi[i, 3]]] = Loewe_ref
  Loewe_CI_list [[listofcombi[i, 3]]] = Loewe_CI
  Excess_loewe_list [[listofcombi[i, 3]]] = Excess_loewe
  
  setTxtProgressBar(pb, i)
}
options(warn = 0)
close(pb)

Loewe_ref_list <- rbindlist(Loewe_ref_list, idcol=TRUE) 
colnames(Loewe_ref_list) <- c("Combination", "Sample_ID",paste0("Conc_", 1:7))
Loewe_ref_list <- inner_join(Loewe_ref_list, HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))

Loewe_CI_list <- rbindlist(Loewe_CI_list, idcol=TRUE)
colnames(Loewe_CI_list) <- c("Combination", "Sample_ID",paste0("Conc_", 1:7))
Loewe_CI_list <- inner_join(Loewe_CI_list, HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
Loewe_CI_list$CI = apply(Loewe_CI_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))

Excess_loewe_list <- rbindlist(Excess_loewe_list, idcol=TRUE)
colnames(Excess_loewe_list) <- c("Combination", "Sample_ID",paste0("Conc_", 1:7))
Excess_loewe_list <- inner_join(Excess_loewe_list, HSA_eval[,c(1,4,5)], by=c("Combination", "Sample_ID"))
Excess_loewe_list$excess_loewe = apply(Excess_loewe_list[,3:9], MARGIN = 1, FUN = function(x) median(x, na.rm=T))

save(Loewe_ref_list, file="output/supp/Loewe_ref_list.RData")
save(Loewe_CI_list, file="output/supp/Loewe_CI_list.RData")
save(Excess_loewe_list, file="output/supp/Excess_loewe_list.RData")

```