---
title: "Clustering and gene signatures (Fig5a-d; SFig6a-d)"
description: "unbiased clustering of top 3000 variably expressed genes for 341 samples"
date: "30 dec 25"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

- refer to src/preprocess/4.rnaseq_preprocess.R for clustering
# runs on the older version to obtain the original clusters
```{r}
# previous versions
# package.dir="package/"
# renv::install(paste0(package.dir, "seurat-4.3.0.tar.gz"))
# renv::install(paste0(package.dir, "seurat-object-4.1.3.tar.gz"))
# renv::install(paste0(package.dir, "harmony_1.0.1.tar.gz"))
# renv::install(paste0(package.dir, "Matrix_1.6-1.1.tar.gz"))
# renv::install(paste0(package.dir, "irlba_2.3.5.1.tar.gz"))

```

# Loading libraries
<https://cran.r-project.org/src/contrib/Archive/harmony/>
<https://github.com/satijalab/seurat/releases>
<https://github.com/satijalab/seurat-object/releases>
```{r librariesAnalysis, echo = TRUE, message = FALSE, warning = FALSE}
library(data.table)
library(Matrix) # Matrix_1.6-1
library(patchwork)
library(rstatix)

library(Seurat) # Seurat_4.3.0.1
library(SeuratObject) # seurat-object-4.1.3
library(harmony) ## must use 1.0.1 version

set.seed(10)

library(tidyverse)
source("src/function/functions.R")
source("src/function/loading_data.R")
```

# Loading data
- 341 oshu samples with RNA
- columns: mutations 34:75, single drugs 76:122, combi 123:269

```{r}
raw_counts <- read.csv("data/oshu/rnaseq_rawcount.csv", check.names = F) # grch37
rownames(raw_counts) <- raw_counts$GeneID
gene <- raw_counts[,1:2]

sampleinfo2 <- read.csv("data/oshu/rnaseq_sampleinfo.csv", check.names = F) 
rownames(sampleinfo2) <- sampleinfo2$Sample_ID
sampleinfo2$cohort <- factor(sampleinfo2$cohort)

raw_counts <- raw_counts[,sampleinfo2$Sample_ID]
```

# clustering
```{r}
## PARAMETERS
NFEATURES = 3000
n=30 #npcs
```

## creating seurat object
```{r}
oshu <- CreateSeuratObject(counts = raw_counts, project = "Sample_ID", min.cells = 3, min.features = 200, meta.data = sampleinfo2, assay="RNA")

oshu <- NormalizeData(oshu)
oshu <- FindVariableFeatures(oshu, selection.method = "vst", nfeatures = NFEATURES)
all.genes <- rownames(oshu)
oshu <- ScaleData(oshu, features = all.genes, min.cells.to.block = 341)
```

## accounting for cohort difference
```{r}
oshu <- RunPCA(oshu, features = VariableFeatures(object = oshu),
               assay = "RNA", seed.use = 42,
               npcs = n, verbose = F)

oshu <- RunHarmony(oshu, verbose = F, 
				group.by.vars = c("cohort"),
				reduction.use = "pca", assay.use = "RNA", reduction.save = "harmony")

DimPlot(oshu, reduction = "pca", group.by = "cohort", pt.size = 1) +
  stat_ellipse(aes(x = PC_1, y = PC_2, group = cohort, color = cohort), type = "norm")

DimPlot(oshu, reduction = "harmony", group.by = "cohort", pt.size = 1) +
  stat_ellipse(aes(x = harmony_1, y = harmony_2, group = cohort, color = cohort), type = "norm")
# ggsave("plot/pca_ac.png", units="in", width=4, height=3, dpi=300)
```

## cluster the samples
To make sure our Harmony integration is reflected in the data visualization, we still need to generate a UMAP derived from these harmony embeddings instead of PCs:
```{r message=FALSE, warning=FALSE}
oshu <- FindNeighbors(oshu, dims = 1:n, k.param = 20, reduction = "harmony")

oshu <- FindClusters(oshu, algorithm = 4, random.seed = 0, verbose = F, resolution = 0.6)

new.cluster.ids <- paste0("Cluster ", levels(oshu$RNA_snn_res.0.6), "\n(n=", table(oshu$RNA_snn_res.0.6), ")")
levels(oshu$RNA_snn_res.0.6) <- new.cluster.ids

oshu <- RunUMAP(oshu, dims = 1:n, n.components = 2L, n.neighbors = 30, seed.use = 86, verbose = F, reduction = "harmony")

p <- DimPlot(oshu, reduction = "umap", group.by = "RNA_snn_res.0.6", pt.size = 2) +
  labs(title = NULL) + theme_classic(base_size = 16) +
  theme(legend.position = "none")
LabelClusters(p, id = "RNA_snn_res.0.6",  fontface = "bold", color = "black", size = 5, bg.color = "white", bg.r = .15, repel = T)
```

## gene signature scores: inflammation
<https://www.waltermuskovic.com/2021/04/15/seurat-s-addmodulescore-function/>

- Warning: The following features are not present in the object: ENSG00000211675, ENSG00000211592, ENSG00000271503, ENSG00000261371, ENSG00000277632, ENSG00000277796, not searching for symbol synonyms (6)
IGLC1, IGKC, CCL5, PECAM1, CCL3, CCL3L1
```{r}
## hngc gene symbols
genes <- read.csv("data/baml/fullgene_baml_75vs75vs90_hngc.csv")
genes <- genes[genes$gene_id %in% gene$GeneID,]
rownames(genes) <- seq(1:nrow(genes))

## getting corr ENSG from Symbol
reference <- read.csv("data/inflammatory_sig_246.csv")
inflammatory_sig_genes <- hgnc.index(reference$gene_id, genes)
inflammatory_sig_genes <- genes[inflammatory_sig_genes$index,c("gene_id", "gene_name.x")] %>% na.omit(.) # 240
colnames(inflammatory_sig_genes) <- c("GeneID", "Gene")

saveRDS(inflammatory_sig_genes, "data/inflammatory_sig_genes.rds")

oshu <- AddModuleScore(object = oshu, features = list(inflammatory_sig_genes$GeneID))
colnames(oshu@meta.data)[ncol(oshu@meta.data)] <- "inflammatory_sig_genes"
```

## tbl 12: saving
```{r message=FALSE, warning=FALSE}
saveRDS(oshu, file="output/rds/oshu_seurat_obj.rds")

cluster <- sampleinfo2 %>%
  dplyr::select(Sample_ID,patientId,dbgap_subject_id,dbgap_rnaseq_sample,dbgap_dnaseq_sample, sampleid, cohort) %>%
  mutate(cluster= oshu@active.ident)
saveRDS(cluster, file="output/rds/oshu_cluster_info.rds")

mat <- t(oshu@meta.data[,c(36:47,50:77)])
top15 <- names(rev(sort(rowSums(mat, na.rm = T))))[1:15]
oshu@meta.data %>%
  select(4, 279, 5, 7, 13:26, top15, 271:277, 280) %>% 
  write.csv("output/oshu_rnaseq_sampleinfo.csv", row.names = F)
```

# tbl 13 - refer to 7.clusterassoc.Rmd


# tbl 14: differential gene expression
```{r}
roc.markers <- FindAllMarkers(oshu, only.pos = FALSE, min.pct = 0.5, logfc.threshold = -Inf, random.seed = 1, test.use = 'roc', return.thresh = 0, verbose=T) %>%
  left_join(., gene, by=c("gene"="GeneID"))

write.csv(roc.markers, "output/oshu_rnaseq_markers_roc.csv", row.names = F)
```

# tbl 15: gene set enrichment analysis: 
- min.diff.pct -> Pre-filter features whose detection percentages across the two groups are similar (within 0.25)
- <https://nbisweden.github.io/workshop-scRNAseq/labs/compiled/seurat/seurat_05_dge.html#Gene_Set_Enrichment_Analysis_(GSEA)>
- obtaining the full gene list between cluster and the rest for GSEA - no arbitrary definition in what is significant gene

```{r message=FALSE, warning=FALSE}
MARKER <- read.csv("output/oshu_rnaseq_markers_roc.csv") %>%
  mutate(rank.col = sign(avg_log2FC) * power)

## loading database for GSEA
hall_set <- load_gs(category = "H", genes = genes$gene_id)

x <- lapply(1:6, function(i){
  tmp <- MARKER[MARKER$cluster==i,] 
  
  original_gene_list <- tmp[["rank.col"]] 
  names(original_gene_list) <- tmp[["gene"]]
  gene_list<-na.omit(original_gene_list)
  gene_list = sort(gene_list, decreasing = TRUE)
  
  gse <- clusterProfiler::GSEA(gene_list, TERM2GENE = hall_set[,c("gs_name", "ensembl_gene")], seed = TRUE, verbose = F, minGSSize=10,  pvalueCutoff = 0.25)
  
  gse@result$Description <- str_replace_all(gse@result$Description, c("^[^_]*_"="", "_"=" "))
  
  x <- as.data.frame(unique(subset(hall_set, select=-c(ensembl_gene, gene_symbol))) )
  gse@result <- left_join(gse@result, x, by= c("ID"="gs_name"))
  
  gse@result
})

df_long = data.frame()
for (i in 1:6){
  df <- x[[i]] %>% # [,c(1,2,5,6,7)]
    mutate(cluster = i, sign_NES = ifelse(sign(NES)==1, "Pos", "Neg"))
  df_long <- rbind(df_long, df)
}


## gene id to symbol
id2symbol <- setNames(gene$Gene, gene$GeneID)
df_long <- df_long %>%
  mutate(symbols = sapply(core_enrichment, function(x) {
    ids <- str_split(x, "/", simplify = FALSE)[[1]]
    syms <- id2symbol[ids]
    out <- paste(syms, collapse = "/")
    if (str_ends(x, "/")) out <- paste0(out, "/")
    out
  }))

write.csv(df_long, "output/oshu_GSEA_hall_cluster.csv", row.names = F)
```

```{r message=FALSE, warning=FALSE}
## loading database for GSEA
go_set <- load_gs(category = "C5", subcategory = "GO:BP", genes = gene$GeneID) 

x <- lapply(1:6, function(i){
  tmp <- MARKER[MARKER$cluster==i,] 
  
  original_gene_list <- tmp[["rank.col"]] 
  names(original_gene_list) <- tmp[["gene"]]
  gene_list<-na.omit(original_gene_list)
  gene_list = sort(gene_list, decreasing = TRUE)
  
  gse <- clusterProfiler::GSEA(gene_list, TERM2GENE = go_set[,c("gs_name", "ensembl_gene")], seed = TRUE, verbose = F, minGSSize=10,  pvalueCutoff = 0.05)
  
  gse@result$Description <- str_replace_all(gse@result$Description, c("^[^_]*_"="", "_"=" "))
  
  x <- as.data.frame(unique(subset(go_set, select=-c(ensembl_gene, gene_symbol))) )
  gse@result <- left_join(gse@result, x, by= c("ID"="gs_name"))
  
  gse@result
})

df_long2 = data.frame()
for (i in 1:6){
  df <- x[[i]] %>% # [,c(1,2,5,6,7)]
    mutate(cluster = i, sign_NES = ifelse(sign(NES)==1, "Pos", "Neg"))
  df_long2 <- rbind(df_long2, df)
}

## gene id to symbol
df_long2 <- df_long2 %>%
  mutate(symbols = sapply(core_enrichment, function(x) {
    ids <- str_split(x, "/", simplify = FALSE)[[1]]
    syms <- id2symbol[ids]
    out <- paste(syms, collapse = "/")
    if (str_ends(x, "/")) out <- paste0(out, "/")
    out
  }))

write.csv(df_long2, "output/oshu_GSEA_gobp_cluster.csv", row.names = F)

```

```{r}
x <- bind_rows(
  df_long %>% mutate(geneset = "HALL"),
  df_long2 %>% mutate(geneset = "GOBP")
)

write.csv(x, "output/oshu_GSEA_hall_gobp_cluster.csv", row.names = F)
```

# plots
```{r message=FALSE}
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(ggpubr)
library(ggrepel)
library(survival)
library(survminer)

source("src/function/colour.R")

oshu <- readRDS("output/rds/oshu_seurat_obj.rds")
df_long <- read.csv("output/oshu_GSEA_hall_cluster.csv")
```

# fig 5
## 5a: umap
```{r}
p <- DimPlot(oshu, reduction = "umap", group.by = "RNA_snn_res.0.6", pt.size = 2) + 
  labs(title = NULL) + theme_classic(base_size = 16) +
  theme(legend.position = "none") + theme_pdf
p <- LabelClusters(p, id = "RNA_snn_res.0.6",  fontface = "bold", color = "black", size = 5, bg.color = "white", bg.r = .15, repel = T) & theme_pdf

ggsave(plot = p, "plot/fig5/main_5a_umap.png", units="in", width=7, height=5, dpi=600, bg = "transparent")
ggsave(plot = p, "plot/fig5/main_5a_umap.pdf", units="in", width=7, height=5, bg = "transparent", device = cairo_pdf)
```

## 5b: heatmap
```{r}
oshu_SI <- oshu@meta.data %>%
  mutate(cluster = seurat_clusters,
         AMLFusion = factor(outcome, levels=fusion.order),
         Disease.status = factor(Disease.status, levels=disease.order3, labels= c("Diagnosis", "Relapse", "Residual", "PostChemo", "Unknown")),
         ELN2022 = factor(ELN2022, levels=eln.col.order)) %>%
  arrange(Disease.status, ELN2022, FAB_subtype)

# Mutation matrix
mat <- t(oshu_SI[,c(36:47,50:77)])
top15 <- names(rev(sort(rowSums(mat, na.rm = T))))[1:15]
mat <- mat[top15,]
colnames(mat) <- oshu_SI$Sample_ID

na_mask <- data.frame(col = 1:341,
                      na = colSums(is.na(mat)),
                      count = colSums(mat, na.rm = T) ) %>%
  filter(na>0) %>%
  arrange(na, count)
na_mask <- na_mask$col

clusters <- oshu@active.ident
color_list <- ggplotColours(n=6)

col_ha = columnAnnotation(
  Cluster=oshu_SI$cluster,
  Disease_status = oshu_SI$Disease.status,
  FAB_subtype=oshu_SI$FAB_subtype,
  ELN2022=oshu_SI$ELN2022,
  AMLFusion=oshu_SI$AMLFusion,
  col=list(Cluster=setNames(color_list, c(1:6)),
           Disease_status = disease.col3,
           FAB_subtype=subtype.col2,
           ELN2022=eln.col,
           AMLFusion=fusion
           ),
  na_col = "grey65",
  annotation_name_side = "left")

# OncoPrint for top mutations
mat2=mat
mat2[is.na(mat2)|mat2==0] = ""
mat2[mat2=="TRUE"] = "MUT;"
col = c("MUT" = "darkblue")
alter_fun = list(
    background = alter_graphic("rect", fill = "grey90",
                               horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),   
    MUT = alter_graphic("rect", fill = col["MUT"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt"))
)
heatmap_legend_param = list(title = "Mutation", at = c("MUT"), 
        labels = c("Mutation"))

op <- oncoPrint(mat2, pct_gp = gpar(fontsize = 11),
    alter_fun = alter_fun, col = col, 
    remove_empty_columns = FALSE, remove_empty_rows = TRUE,
    pct_side = "right", row_names_side = "left") 

x <- ComplexHeatmap::draw(op)
col_dend = column_order(x)
col_dend = c(col_dend[!col_dend %in% na_mask], setNames(na_mask,na_mask))

#########################################
mat[is.na(mat)] = "NA;"
mat[mat=="FALSE"] = "NEG;"
mat[mat=="TRUE"] = "MUT;"
col = c("MUT" = "darkblue",
        "NEG" = "grey90",
        "NA" = "grey75")
alter_fun = list(
    background = alter_graphic("rect", fill = "grey75", horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),   
    MUT = alter_graphic("rect", fill = col["MUT"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),
    NEG = alter_graphic("rect", fill = col["NEG"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt")),
    `NA` = alter_graphic("rect", fill = col["NA"], horiz_margin = unit(.2, "pt"), vertical_margin = unit(.2, "pt"))
)

heatmap_legend_param = list(title = "Mutation", at = c("MUT", "NEG", "NA"), 
        labels = c("Present", "Absent", "NA"))

op <- oncoPrint(mat, 
    column_names_gp = gpar(fontsize = 12),pct_gp = gpar(fontsize = 12),
    alter_fun = alter_fun, col = col, column_order = col_dend, column_split = oshu_SI$cluster,
    remove_empty_columns = FALSE, remove_empty_rows = TRUE,
    pct_side = "right", row_names_side = "left", bottom_annotation = col_ha,               
    heatmap_legend_param = heatmap_legend_param, 
    top_annotation = HeatmapAnnotation(cbar = anno_oncoprint_barplot(type = c("MUT"))),
    right_annotation = rowAnnotation(rbar = anno_oncoprint_barplot(type = c("MUT"))),
    pct_include = c("MUT"),
    height = unit(0.5 * nrow(mat), "cm")
    )

# Cell state heatmap
mat <- t(oshu_SI[,c("LSPC.Quiescent", "LSPC.Primed.Top100", "LSPC.Cycle.Top100", "GMP.like.Top100", "ProMono.like.Top100", "Mono.like.Top100", "cDC.like.Top100")])
hm <- Heatmap(mat, rect_gp = gpar(col = "white", lwd = 0.2), na_col = "grey90",
    col = colorRamp2(c(-4, -1, 0, 1, 4), c("darkblue", "blue", "white","red", "darkred")), 
    row_names_side = "left", show_column_names = F, cluster_rows = F, cluster_columns = F,
    row_labels = c("LSPC-Quiescent", "LSPC-Primed", "LSPC-Cycle", "GMP-like", "ProMono-like", "Mono-like", "cDC-like"),
    column_split = oshu_SI$cluster, heatmap_legend_param = list(title = "z score"),
    height = unit(3.5, "cm"))

# Percent heatmap
library(viridis)
mat <- t(oshu_SI[,c(19:26)])
rownames(mat) <- gsub("percent","%", rownames(mat))
hm1 <- Heatmap(mat, rect_gp = gpar(col = "white", lwd = 0.2),
    col = viridis(10), na_col = "grey75", 
    row_names_side = "left", show_column_names = F, cluster_rows = F, cluster_columns = F,
    column_split = oshu_SI$cluster, heatmap_legend_param = list(title = "Percent"),
    height = unit(4, "cm"))

ht_list = op  %v% hm1 %v% hm

# png("plot/fig5/main_5b_hm_overview_rna.png", width=45, height=27, units="cm", res=600, bg = "transparent", family = "Arial")
Cairo::CairoPDF("plot/fig5/main_5b_hm_overview_rna.pdf", width=45/2.54, height=27/2.54, bg = "transparent", family = "Arial")
ComplexHeatmap::draw(ht_list, heatmap_legend_side = "bottom", annotation_legend_side = "bottom", merge_legend=T, background = "transparent")
dev.off()
```

## 5c: bubble plot
```{r}
p <- df_long %>%
  filter(p.adjust < 0.05) %>%
  group_by(ID) %>%
  mutate(n = n(), Description=str_trunc(Description, 95),
         cluster = factor(cluster, levels=c(1,3,5,2,4,6)),
         order = median(p.adjust)
         ) %>%
  filter(n>4) %>%
ggplot(., aes(x = cluster, y = reorder(Description, -order), size = -log10(p.adjust), fill = NES)) + 
  geom_point(shape = 21) +
  scale_y_discrete(labels = label_wrap_gen(50)) +
  scale_size_continuous(range = c(1, 8), breaks = c(0, 1, 2, 3)) +
  scale_fill_gradientn(colours = pals::coolwarm(100)) +
  ylab("Hallmark Pathway") +
  xlab("Clusters") +
  theme_classic(base_size = 12) +
  theme(axis.text = element_text(color="black")) +
  theme_pdf

ggsave(plot = p, "plot/fig5/main_5c_bubbleplot_gsea_hall.png", units="in", width=7, height=7, dpi=600, bg = "transparent") 
ggsave(plot = p, "plot/fig5/main_5c_bubbleplot_gsea_hall.pdf", units="in", width=7, height=7, bg = "transparent", device = cairo_pdf) 
```

## 5d: umap & box

```{r}
## umap
p <- FeaturePlot(oshu, reduction = "umap", features = "inflammatory_sig_genes", pt.size = 2, order=TRUE)  + 
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu")), name = "Inflammatory\nScore") + 
  ggtitle("") +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.8, 0.8), 
        legend.text = element_text(size=10), 
        legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size=12), 
        aspect.ratio = 0.7) 
p <- p & theme_pdf

ggsave(plot = p, "plot/fig5/main_5d_umap_inflam.png", units="in", width=7, height=5, dpi=600, bg = "transparent")
ggsave(plot = p, "plot/fig5/main_5d_umap_inflam.pdf", units="in", width=7, height=5, bg = "transparent", device = cairo_pdf)
```

```{r}
## boxplot of inflammatory score against clusters with KW
x <- data.frame(cluster = oshu@meta.data$seurat_clusters, score = oshu@meta.data$inflammatory_sig_genes)

p <- ggboxplot(x, x = "cluster", y = "score", fill = "cluster", 
        legend ="none", alpha=0.7,
        xlab="Clusters", ylab="Inflammatory Score", 
        add = "jitter", add.params = list(alpha = 0.6)) +
  scale_fill_manual(values=color_list) + 
  stat_compare_means(method="kruskal", label.x = 1.3) + 
  theme(aspect.ratio = 1.4) +
  theme_pdf

ggsave(plot = p, "plot/fig5/main_5d_boxplot_inflam.png", units="in", width=5, height=4.8, dpi=600, bg = "transparent")
ggsave(plot = p, "plot/fig5/main_5d_boxplot_inflam.pdf", units="in", width=5, height=4.8, bg = "transparent", device = cairo_pdf)

```

# supp fig 6
## 6a: kaplan meier survival curve
```{r}
# using unique patients; taking the sample with the earlier point
df <- oshu@meta.data[,c("Sample_ID","patientId", "Disease.status", "overallSurvival", "isDead", "seurat_clusters")] %>%
  arrange(Sample_ID) %>%
  group_by(patientId) %>%
  slice_head(n=1)

m.fit <- survfit(Surv(time=overallSurvival, event=isDead, type="right") ~  seurat_clusters, data = df)

m.surv <- ggsurvplot(m.fit, data = df, pval=F, 
                     break.time.by = 250,
                     ggtheme=theme_classic2(),
                     legend.labs = paste("Cluster", 1:6),
                     risk.table = T,
                     legend.title = "", xlab = "Time (Day)",
                     log.rank.weights = "1" #log-rank test
                     ) 

p <- m.surv[[1]] / m.surv[[2]] + plot_layout(heights = c(0.8,0.2)) & theme_pdf

ggsave(plot = p, "plot/suppfig6/supp_6a_survival.png", units="cm", width=20, height=20, dpi = 600, bg = "transparent")
# ggsave("plot/suppfig6/supp_6a_survival.svg", units="cm", width=20, height=20)
ggsave(plot = p, "plot/suppfig6/supp_6a_survival.pdf", units="cm", width=20, height=20, bg = "transparent", device = cairo_pdf)

pairwise_survdiff(Surv(time=overallSurvival, event=isDead, type="right") ~  seurat_clusters, data = df, p.adjust.method = "none")
```

	Pairwise comparisons using Log-Rank test 

data:  df and seurat_clusters 

  1     2     3     4     5    
2 0.210 -     -     -     -    
3 0.267 0.594 -     -     -    
4 0.980 0.302 0.358 -     -    
5 0.028 0.301 0.102 0.065 -    
6 0.451 0.083 0.160 0.378 0.044

P value adjustment method: none 

## 6b: pointrange, gene sig
```{r}
gene_sig2  <- oshu@meta.data[,c("LSPC.Quiescent", "LSPC.Primed.Top100", "LSPC.Cycle.Top100", "GMP.like.Top100", "ProMono.like.Top100", "Mono.like.Top100", "cDC.like.Top100", "seurat_clusters")] %>%
  pivot_longer(cols = 1:7, names_to = "Cell.Type") %>%
  mutate(
    Cell.Type = factor(
      Cell.Type, levels=(c("LSPC.Quiescent", "LSPC.Primed.Top100", "LSPC.Cycle.Top100", "GMP.like.Top100", "ProMono.like.Top100", "Mono.like.Top100", "cDC.like.Top100")), 
      labels=(c("cDC-like", "Mono-like", "ProMono-like", "GMP-like", "LSPC-Cycle", "LSPC-Primed", "LSPC-Quiescent"))
      ),
    clusters = factor(seurat_clusters, levels=c(6,5,4,3,2,1))
  )

p <- ggplot(gene_sig2, aes(x=Cell.Type, y=value, color=clusters)) +
  scale_color_manual(values=setNames(color_list, 1:6), breaks = c(1:6), name = "Cluster") +
  geom_hline(yintercept = 0) +
  stat_summary(fun.data = 'mean_sdl', #orientation = "y", 
               geom = "pointrange", position=position_dodge(width=0.5))  +
  labs(y="Standardised Gene Signature Score", x="") + 
  theme_classic(base_size = 15) + 
  theme(aspect.ratio = 1.4,
        legend.position = "top", 
        axis.text.y = element_text(colour = c(rep("navy",3), "black", rep("firebrick",3)))
        ) + 
  stat_compare_means(
    aes(x=Cell.Type, y=value, group=clusters), 
    method = "kruskal.test", 
    label = "p.signif", coord.flip = TRUE, inherit.aes = F) +
  coord_flip() +
  theme_pdf

ggsave(plot = p, "plot/suppfig6/supp_6b_pointrange_genesig.png", units="in", width=6, height=8, dpi=600)
ggsave(plot = p, "plot/suppfig6/supp_6b_pointrange_genesig.pdf", units="in", width=6, height=8, bg = "transparent", device = cairo_pdf)
```

## 6c: DEG

```{r}
plot <- read.csv("output/oshu_rnaseq_markers_roc.csv")
log2fc_thres = 0.25
power_thres = 0.4

# Preprocess data
plot <- plot %>%
  mutate(
    sig = ifelse(power > 0.5, T, F),
    signed_power = sign(avg_log2FC) * power
  ) %>%
  group_by(cluster, sig, sign(avg_log2FC)) %>%
  mutate(
    rank = rank(-abs(avg_log2FC)),
    label = if_else(grepl("RP", Gene), NA_character_, Gene),
    label_up = if_else(sig & rank <= 10 & sign(avg_log2FC) == 1, label, NA_character_),
    label_down = if_else(sig & rank <= 10 & sign(avg_log2FC) == -1, label, NA_character_)
  ) %>%
  arrange(signed_power) %>% select(-label)

# Jitter position for points
pos <- position_jitter(width = 0.3, height = 0, seed = 2)

# Create plot
p <- ggplot(plot, aes(x = cluster, y = avg_log2FC)) +
  geom_point(aes(color = signed_power, alpha = power > 0.5), 
             position = pos, stroke = 0.3, size = 2) +
  scale_color_gradientn(colors = c("#3B4CC0", "grey", "#B40426"), values = c(0, 0.3, 0.7, 1)) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.3)) +

  # Label top up- and down-regulated genes
  geom_text_repel(aes(label = label_up), position = pos, seed = 123,
                  max.overlaps = 20, color = "black", size = 8 / ggplot2:::.pt,
                  ylim = c(0.2, 3.6), direction = "both", box.padding = 0.5,
                  point.padding = 0.25, min.segment.length = 0) +
  geom_text_repel(aes(label = label_down), position = pos, seed = 123,
                  max.overlaps = 20, color = "black", size = 8 / ggplot2:::.pt,
                  ylim = c(-3.6, -0.2), direction = "both", box.padding = 0.5,
                  point.padding = 0.25, min.segment.length = 0) +

  # Add horizontal lines and cluster labels
  geom_hline(yintercept = c(-0.58, -0.25, 0, 0.25, 0.58), 
             linetype = c("solid", "dashed", "solid", "dashed", "solid")) +
  geom_label(data = data.frame(x = 1:6, label = paste("Cluster", 1:6)),
             aes(x = x, y = 0, label = label), fill = "white",
             size = 10 / ggplot2:::.pt, alpha = 1) +

  # Theming
  theme_pubr() +
  labs(y = "Average Log2 Fold Change", x = NULL) +
  ylim(-3, 3.6) +
  theme(
    aspect.ratio = 0.6,
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.text = element_text(size = 10),
    text = element_text(size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 10)
  ) + theme_pdf

ggsave(plot = p, "plot/suppfig6/supp_6c_cluster_deg.png", units="in", width=11, height=8, dpi=600, bg = "transparent")
ggsave(plot = p, "plot/suppfig6/supp_6c_cluster_deg.pdf", units="in", width=11, height=8, bg = "transparent", device = cairo_pdf)
```

## 6d: trend curves
```{r}
df <- oshu@meta.data[,c("Sample_ID", "seurat_clusters", "inflammatory_sig_genes", "LSPC.Quiescent", "LSPC.Primed.Top100", "LSPC.Cycle.Top100", "GMP.like.Top100", "ProMono.like.Top100", "Mono.like.Top100", "cDC.like.Top100")] %>%
  pivot_longer(cols=4:10, names_to = "cell_type", values_to = "cell_type_score") %>%
  mutate(cell_type = factor(cell_type, levels=cell.type.order, labels=cell.type.name)) 


## smooth line plot of cell maturation score against inflammatory score 
p <- ggplot(df, aes(x=inflammatory_sig_genes, y=cell_type_score, color=cell_type)) +
  stat_smooth(formula = y ~ s(x, bs = "cs"), method = "gam", se = F) +
  labs(x="Inflammatory Score", y="Cell Maturation Score", color="Cell Maturation Type") +
  scale_color_manual(values = hierarchy_colors) +
  theme_pubr() + theme(aspect.ratio = 1, legend.position = "right") + theme_pdf

ggsave(plot = p, "plot/suppfig6/supp_6d_trend_inflam_mature.png", units="in", width=6, height=4, dpi=600)
ggsave(plot = p, "plot/suppfig6/supp_6d_trend_inflam_mature.pdf", units="in", width=6, height=4, bg = "transparent", device = cairo_pdf)
```

# session info
```{r}
sessionInfo()
```