---
title: "Gene Co-expression Network (Fig5e; SFig6e,f)"
description: "using VST top 3000 variably expressed genes for 341 samples"
date: "30 dec 25"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

# references
<https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0>
<https://alexslemonade.github.io/refinebio-examples/04-advanced-topics/network-analysis_rnaseq_01_wgcna.html>
<https://moodle.france-bioinformatique.fr/pluginfile.php/703/mod_folder/content/0/WGCNA/WGCNA.html?forcedownload=1>

GWENA vignette: <https://bioconductor.org/packages/devel/bioc/vignettes/GWENA/inst/doc/GWENA_guide.html>

- tried GWENA and CEMiTools

# Loading libraries
```{r librariesAnalysis, echo=TRUE, message=FALSE, warning=FALSE}
library(WGCNA)
library(Seurat)
library(data.table)
library(tidyverse)
library(CEMiTool)
set.seed(10)
source("src/function/functions.R")
source("src/function/colour.R")
source("src/function/loading_data.R")
```

# loading data
```{r}
cluster <- readRDS(file="output/rds/oshu_cluster_info.rds")
oshu <- readRDS("output/rds/oshu_seurat_obj.rds")
selected_genes <- Seurat::VariableFeatures(oshu) # top 3000 HVG from seurat

expr_normalized <- readRDS("data/oshu/rnaseq_vst.rds")
rownames(expr_normalized) <- expr_normalized$GeneID
gene <- expr_normalized[,1:2]
```

```{r}
metadata <- oshu@meta.data[,c(5,7,13:280)]
expr_normalized <- expr_normalized[selected_genes, rownames(metadata)] 
selected_genes <- gene[selected_genes,]

rownames(expr_normalized) <- selected_genes$Gene

metadata$ID <- colnames(expr_normalized)
metadata$cluster <- cluster$cluster

go_set <- load_gs(category = "C5", subcategory = "GO:BP", genes = gene$GeneID) %>%
  mutate(gs_name = gsub("_", " ", gs_name))
colnames(go_set)[c(1,3)] = c("term", "gene")
hall_set <- load_gs(category = "H", genes =  gene$GeneID) %>%
  mutate(gs_name = gsub("_", " ", gs_name))
colnames(hall_set)[c(1,3)] = c("term", "gene")

gs <- hall_set[,c(1,3)]
```

# Running CEMiTool ( a wrapper for WGCNA )
<https://github.com/csbl-usp/CEMiTool/blob/master/R/modules.R>

- using obtaining VST normalised matrix
- identifying gene modules across all the counts

```{r}
library(doParallel)
cl <- makeCluster(4) # Use available cores minus one
registerDoParallel(cl)

cem <- cemitool(expr = expr_normalized,
                annot = metadata,
                sample_name_column = "ID",
                class_column = "cluster",
                filter = F, diss_thresh = 0.75,
                gmt = gs,
                cor_method = "spearman",
                verbose = T
                )

saveRDS(cem, "output/rds/oshu_wgcna.rds")

# write analysis results into files
# write_files(cem, directory="output/cemi/", force=T)
```


# downstream analysis
```{r message=FALSE}
library(ComplexHeatmap)
library(clusterProfiler)
library(enrichplot)

# run # loading data
cem <- readRDS("output/rds/oshu_wgcna.rds")

# extracting gene modules
genes <- module_genes(cem)
genes$modules <- gsub("M", "Mod", genes$modules)

ordering <- c(paste0("Mod", 1:(length(unique(genes$modules))-1) ), "Not.Correlated")
col2 <- setNames(c(Custom[1:(length(ordering)-1)], "black"), ordering)

# loading genesets
reference <- read.csv("data/inflammatory_sig_246.csv")
andyzeng.gs <- load_genesets("data/AMLCellType_Genesets_all.gmt", ignore_cols = 2) 
```

## tbl 16: saving genes
Adjacency Method:

This method uses the adjacency matrix to evaluate the connectivity of genes within a module.
Genes are ranked based on the row sums of their adjacency values, which represent the strength of their connections to other genes in the module.
The top genes with the highest connectivity are selected as hub genes.

kME Method:

This method uses the module eigengene (ME), which is a representative expression profile for the module.
It calculates the correlation (signed kME) of each gene's expression profile with the module eigengene.
Genes are ranked based on their kME values (correlation scores), and those with the highest values are chosen as hub genes.

```{r}
adj <- get_hubs(cem, n="all", method = "adjacency") # in desc order of connectivity
kme <- get_hubs(cem, n="all", method = "kME") 

hub_genes <- rbindlist(lapply(setNames(names(adj), names(adj)), function(x) 
  data.frame(adj=adj[[x]], 
             kME=kme[[x]][names(adj[[x]])], 
             gene = names(adj[[x]]) ) 
  ), idcol="modules")

## gene membership in other gene sets
genes_df <- hub_genes %>%
  mutate(in_inflammation = ifelse(gene %in% reference$gene_id, T, F),
         modules = gsub("M", "Mod", modules))
temp <- do.call("cbind",lapply(andyzeng.gs, function(x) genes_df$gene %in% x)) %>%
  as.data.frame()
temp$primitivelike <- rowSums(temp[,1:3]) > 0
temp$maturelike <- rowSums(temp[,5:7]) > 0

genes_df <- cbind(genes_df, temp)
write.csv(genes_df, "output/oshu_wgcna_module_genes.csv", row.names = F)
```

## tbl 18: correlation with features
```{r}
MEs <- mod_summary(cem, method='eigengene')
MEs$modules <- gsub("M", "Mod", MEs$modules)

MEs <- MEs %>%
  mutate(modules = factor(modules, levels=ordering)) %>%
  arrange(modules) %>%
  filter(modules != "Not.Correlated") %>% 
  column_to_rownames("modules") 

colnames(MEs) <- rownames(metadata)
MEs <- t(MEs)
nSamples <- nrow(MEs)

temp <- list(
  drug = colnames(metadata)[68:113],
  combi = colnames(metadata)[114:260],
  celltype = colnames(metadata)[c(261:267,270)],
  percentcell =  colnames(metadata)[c(9:16)],
  mutation = colnames(metadata)[26:67],
  prior = colnames(metadata)[17:23],
  cluster = colnames(metadata)[272],
  fab = colnames(metadata)[3],
  fusion = colnames(metadata)[25]
  # status = colnames(metadata)[2]
  )

metadataNum_all <- list()

for (variable in names(temp)) {
  metadataNum <- metadata[, temp[[variable]]]
  
  if (is.null(ncol(metadataNum))){ # to binarize categorical variables with >2 categories
    metadataNum = binarizeCategoricalVariable(metadataNum, includePairwise = FALSE, includeLevelVsAll = TRUE)
  }
  
  metadataNum_all[[variable]] = metadataNum
}
  
metadataNum <- do.call(cbind, metadataNum_all)
moduleTraitCor <- cor(MEs, metadataNum,use = "pairwise.complete.obs")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nSamples)
moduleTraitCor <- reshape2::melt(moduleTraitCor, value.name = "corcoeff")
moduleTraitPvalue <- reshape2::melt(moduleTraitPvalue, value.name = "p.value")
res <- inner_join(moduleTraitCor, moduleTraitPvalue) %>%
  dplyr::rename(module = Var1, trait = Var2) %>%
  separate_wider_delim(cols=trait, delim=".", names = c("trait_category", "trait"), too_many = "merge") %>%
  mutate(trait = ifelse(trait %in% names(listofnames), listofnames[trait], trait))

write.csv(res, file="output/oshu_corr_ME_trait_results.csv", row.names=F)
```

## suppfig 6e: heatmap - cor ME w percentcell
```{r}
moduleTraitCor <- cor(MEs, metadataNum_all[["percentcell"]],use = "pairwise.complete.obs")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nSamples)

## plotting
textMatrix <- paste(signif(moduleTraitCor, 2), "\n(", signif(moduleTraitPvalue, 1), ")", sep = "")
dim(textMatrix) <- dim(moduleTraitCor)
colnames(textMatrix) <- colnames(moduleTraitCor)
rownames(textMatrix) <- rownames(moduleTraitCor)

for(r in row.names(moduleTraitCor)){
  for(c in colnames(moduleTraitCor)){
    if(moduleTraitPvalue[r, c] > 0.05){
      moduleTraitCor[r, c] <- 0
      textMatrix[r, c] <- ""
    }
  }
}

# instead of labeledHeatmap
hm <- Heatmap(moduleTraitCor,
  col = colorRamp2(c(-1, 0, 1),c("darkgreen", "white", "darkmagenta")),
  name = "Pearson_cor", border = T, rect_gp = gpar(col = "grey90"),
  column_names_rot = 45,
  cluster_columns=T, cluster_rows=F,
  row_names_side = "left", row_dend_reorder = T,
  left_annotation = rowAnnotation(
    ME=rownames(moduleTraitCor), 
    col=list(ME=col2), show_legend = F, show_annotation_name = F
    ),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(textMatrix[i, j], x, y, gp = gpar(fontsize = 8))
  },
  width = unit(0.7*ncol(moduleTraitCor), "in"), 
  height = unit(0.5*nrow(moduleTraitCor), "in")
)

# png("plot/suppfig6/supp_6e_hm_wgcna_cell.png", width = (0.7*ncol(moduleTraitCor)) + 3, height = (0.5*nrow(moduleTraitCor)) + 4, units = "in", res=600, bg = "transparent")
Cairo::CairoPDF("plot/suppfig6/supp_6e_hm_wgcna_cell.pdf", width = (0.7*ncol(moduleTraitCor)) + 3, height = (0.5*nrow(moduleTraitCor)) + 4, bg = "transparent", family = "Arial")
ComplexHeatmap::draw(hm, background = "transparent")
```

## tbl 17: annotating with pathways 
- pvalueCutoff (adjusted pvalue cutoff) = 0.05

### hallmark geneset
```{r message=FALSE}
formula_hall <- compareCluster(genes~modules, data=genes, fun="enricher", TERM2GENE = hall_set[,c(1,3)], pvalueCutoff = 0.05) 
formula_hall <- pairwise_termsim(formula_hall)  

formula_hall@compareClusterResult <- formula_hall@compareClusterResult %>%
  mutate(ID = gsub("HALLMARK ", "", ID),
         Description = gsub("HALLMARK ", "", Description)) 

# write.csv(formula_hall@compareClusterResult, paste0("output/oshu_wgcna_ora_hall_q25.csv"), row.names = F)
```

### gobp geneset
- p<0.05 for stricter cut offs
```{r}
formula_go <- compareCluster(genes~modules, data=genes, fun="enricher", TERM2GENE = go_set[,c(1,3)], pvalueCutoff = 0.05)
formula_go@compareClusterResult <- formula_go@compareClusterResult %>%
  filter(Count >1)

df <- bind_rows(formula_hall@compareClusterResult, formula_go@compareClusterResult)

write.csv(df, "output/oshu_wgcna_ora_hall_q25_go_q05.csv", row.names = F)
```

## suppfig 6f: Gene-Concept Network (hallmark geneset)

The cnetplot() depicts the linkages of genes and biological concepts (e.g. GO terms or KEGG pathways) as a network. GSEA result is also supported with only core enriched genes displayed.
```{r}
## cnetplot
options(ggrepel.max.overlaps = 30)
set.seed(1)
p <- cnetplot(formula_hall, node_label = "all", layout = "fr", 
         cex.params=list(category_node=3, gene_node=5, category_label=2, gene_label=1.5),
         seed = 1
         ) + scale_fill_manual(values=col2, name="Module") +
  theme(legend.key.size = unit(0.5, "cm"), legend.text = element_text(size=14)) 


## annotates specific names only
p[["layers"]][[8]][["data"]] <- p[["layers"]][[8]][["data"]] %>%
  mutate(name = ifelse(name %in% reference$gene_id, name, ""))
p <- p + theme_pdf

ggsave(plot = p, "plot/suppfig6/supp_6f_cnetplot.png", width = 15, height = 12, units = "in", dpi = 600, bg = "transparent")
ggsave(plot = p, "plot/suppfig6/supp_6f_cnetplot.pdf", width = 15, height = 12, units = "in", bg = "transparent", device = cairo_pdf)
```

## fig 5e: plotting module summary
- manual annotation
```{r}
row_naming_all <- c(
  "Mod1:INFLAM_RESP",
  "Mod2:IFNG_RESP,TNFA_NFKB,EMT",
  "Mod3:IL2_STAT5",
  "Mod4:TNFA_NFKB",
  "Mod5:HOX_GENES",
  "Mod6:HEME_METABOLISM",
  "Mod7:EMT",
  "Mod8:ER_LATE",
  "Mod9:TCELL_ACTIVATION",
  "Mod10:EMT,COAGULATION",
  "Mod11:BCELL_ACTIVATION"
)
```

```{r}
mod_sum <- mod_summary(cem, method='eigengene')
clust_order <- metadata$cluster

MEs<- mod_sum %>%
  mutate(modules = gsub("M", "Mod", modules),
         modules = factor(modules, levels=ordering)) %>%
  arrange(modules)
rownames(MEs) <- MEs$modules
MEs<- MEs[1:(nrow(MEs)-1),2:ncol(MEs)]

col_ha <- columnAnnotation(
  Cluster = clust_order,
  col = list(Cluster = setNames(color_list, levels(clust_order))),
  annotation_height = unit(0.3, "mm"),
  show_legend=F, annotation_name_side = "left"
)

row_ha <- rowAnnotation(
  Module = rownames(MEs),
  col = list(Module = col2),
  annotation_height = unit(0.3, "mm"), gp = gpar(col = "white", lwd=0.3),
  show_legend=F, show_annotation_name = F
)

hm <- Heatmap(MEs, border = T,border_gp = gpar(col = "black", lwd=2),
        col = circlize::colorRamp2(seq(-0.15, 0.15, length = 11), rev(RColorBrewer::brewer.pal(11, "RdYlBu"))),
        show_row_names = T, show_column_names = F, row_names_side = "right",
        
        row_labels = row_naming_all,
        column_split = clust_order,
        
        show_column_dend = F, show_row_dend = F,
        cluster_columns = T, cluster_rows = F,
        clustering_method_columns = "ward.D2", clustering_method_rows = "ward.D2",
        cluster_column_slices = F, column_title_gp = gpar(fontsize = 12),
        column_dend_reorder = T, row_dend_reorder = T,
        top_annotation = col_ha,
        right_annotation = row_ha,
        row_gap = unit(0.3, "mm"), column_gap = unit(1, "mm"),
        name = "ME",
        width = unit(22, "cm"), height = unit(8, "cm")
        )

# png("plot/fig5/main_5e_hm_wgcna.png", width = 15, height = 5, units = "in", res = 600, bg = "transparent", family = "Arial")
Cairo::CairoPDF("plot/fig5/main_5e_hm_wgcna.pdf", width = 15, height = 5, bg = "transparent", family = "Arial")
ComplexHeatmap::draw(hm, background = "transparent")
```

# session info
```{r}
sessionInfo()
```