---
title: "Single cell - Pseudotime (AML + healthy)"
description: "van galen et al 2019 scRNAseq + pseudotime"
date: "17 October 25"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---
*** refer to src/preprocess/6.scrna_vg_preprocess.Rmd for preprocessing and clustering
- re-clustering using seurat v5 for more efficient processing
- seurat v5 was converted to v3 assay class for seurat 4.3.0 compatibility

# 35 AML samples (N=12) + 5 healthy samples (N=5)
*** currently contains all longitudinal samples

"AML1012" 
"AML210A"  
"AML328"  
"AML329"  
"AML419A" 
"AML420B" 
"AML475"  
"AML556"  
"AML707B" 
"AML870"  
"AML916"  
"AML921A"   
"BM1"     
"BM2"   
"BM3"     
"BM4"     
"BM5"   

To remove as described in van galen et al 2019
"AML314", "AML371", "AML722B", "AML997" 

# library
```{r message=FALSE}
library(tidyverse)
library(Seurat)

library(slingshot)
library(tradeSeq)
# library(SeuratExtend)

library(ComplexHeatmap)
library(RColorBrewer)
library(patchwork)

library(ggpubr)
library(rstatix)

source("src/function/scrna_colour.R")
set.seed(10)

set_alpha <- function(plot, alpha = 0.7){
  for (i in 1:length(plot)) {
    plot[[i]]$layers[[1]]$aes_params$alpha =  alpha
  }
  return(plot)
}
```

# loading overall 
```{r}
all_ <- readRDS(file="data/scrna/all_v3.rds")
reference <- read.csv("data/inflammatory_sig_246.csv")
```

# plots 

## fig 6a: overall umap
```{r}
DimPlot(all_, reduction = "umap", group.by="CellType", shape.by = "cell_status", 
        cols = hierarchy_colors, pt.size=.5, shuffle = T) %>%
  set_alpha() + 
  scale_shape_manual(values = c(17,16,15), name = NULL, 
                     labels = c("AML_malignant\n(n=13,489)", 
                                "AML_nonmalignant\n(n=15,333)", 
                                "Healthy_normal\n(n=6,915)")) +
  theme(aspect.ratio = 1) + ggtitle("") + 
  guides(shape = guide_legend(override.aes = list(size = 3, alpha = 1)),
         color = guide_legend(override.aes = list(size = 3, alpha = 1)) )

ggsave("plot/fig6/main_6a_umap_overall.png", width = 7.5, height = 6, dpi = 600)
```

## suppfig 6a: seurat clusters + batch effects

```{r}
p <- DimPlot(all_, reduction = "umap", group.by = c("PatientID", "seurat_clusters")) + DimPlot(all_, reduction = "umap", group.by = c( "CellType"), cols =  hierarchy_colors, combine = F) 

p %>%
  set_alpha() & 
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)) )

ggsave("plot/suppfig7/supp_7a_afterbc_all_2.png", width=20, height=5)
```

## fig 6b: split by cell status (top)

```{r}
ob_list <- SplitObject(all_, split.by = "cell_status")

p <- lapply(X = ob_list, FUN = function(x) {
  
    DimPlot(x, reduction = "umap", group.by="CellType", split.by = "cell_status", 
            cols = hierarchy_colors, pt.size=.5) %>%
      set_alpha() +
      theme(aspect.ratio = 1, strip.text = element_text(size = 14),
            legend.key.height = unit(0.5, "cm")) + ggtitle("") + 
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)) )
})

p[[2]] + p[[1]] + p[[3]] + patchwork::plot_layout(ncol = 3, guides = "collect")

ggsave("plot/fig6/main_6b_umap_celltype.png", width = 15, height = 5, dpi = 600)
```

## fig 6b: inflammation (middle)

```{r}
p <- FeaturePlot(all_, reduction = "umap", 
                 split.by = "cell_status", features = "Inflammation", 
                 pt.size = .8, order=TRUE, keep.scale = "feature") %>%
  set_alpha()  & 
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "Spectral")),
                         limits=c(-0.15,1), name = "Inflammatory\nScore") & 
  theme(axis.title = element_text(size = 14), aspect.ratio = 1, 
        title = element_text(size = 12), 
        axis.title.y.right = element_blank()
        ) 

p + theme(legend.position = "right")

ggsave("plot/fig6/main_6b_dimplot_infl.png", width = 14, height = 5, dpi = 600)

```

## fig 6b: boxplot of inflammation scores per cell per cell type (bottom)

```{r}
df <- FetchData(all_, vars = c("cell_status", "celltype", "CellType", "Inflammation"))
ggplot(df, aes(x = celltype, y = Inflammation, fill = CellType)) +
  geom_boxplot() +
  scale_fill_manual(values = hierarchy_colors, guide = "none") +
  theme_classic(base_size = 12) +
  labs(x="", y="Inflammatory Score")+
  facet_grid(~cell_status, drop = T, axes = "all") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        aspect.ratio = 0.5)

ggsave("plot/fig6/main_6b_boxplot_infl.png", width = 12, height = 3, dpi = 600)

```

## difference between patients
- annotated based on distribution of each cell subpopulations
  - primitive-like - when HSC, Prog and GMP are most of the malignant cells
  - mature-like - when ProMono, Mono and cDC are most of the malignant cells
- caveat: AML329 - though diagnosis was "monocytic", was annotated as primitive like

```{r}
df <- all_@meta.data %>%
  select(SampleID := orig.ident, PatientID, CellType, cell_status) %>%
  group_by(SampleID, CellType, cell_status, PatientID) %>%
  summarise(n = n()) %>%
  group_by(SampleID, cell_status) %>%
  mutate(total = sum(n),
         PercentDist = (n/total)*100,
         CellType = factor(CellType, levels = rev(unique(CellType))) ) %>% 
  ungroup() %>%
  mutate(Day = gsub(".*-D", "", SampleID))

write.csv(df, "output/supp/scrna/scrna_cell_dist_per_sampleid.csv", row.names = F)

```

## fig 6c: cell distribution bar plot

```{r}
patient_order <- read.csv("data/scrna/vangalen_sampleinfo.csv") %>%
  filter(Sample %in% unique(all_$PatientID)) %>%
  select(SampleID, Sample, Day := Days.from.Diagnosis, AML_subtype, Category) %>%
  slice_min(order_by = Day, n=1) %>% arrange(desc(Category), SampleID)

df <- read.csv("output/supp/scrna/scrna_cell_dist_per_sampleid.csv", check.names = F)
df_1 <- df %>%
  mutate(
    category = case_when(
      CellType %in% c("HSC-like", "Prog-like", "HSC", "Prog") ~ "primitive",
      CellType %in% c("ProMono-like", "Mono-like", "cDC-like", "ProMono", "Mono", "cDC") ~ "mature",
      CellType %in% c("GMP-like", "GMP") ~ "GMP",
      CellType %in% c('ProB', 'B', 'Plasma', 'T', 'CTL', 'NK') ~ "lymphoid",
      CellType %in% c('earlyEry', 'lateEry') ~ "erythroid",
      .default = CellType
    )
  ) %>% 
  group_by(SampleID, category, cell_status, total, Day) %>%
  summarise(n = sum(n)) %>%
  group_by(SampleID, cell_status) %>%
  mutate(PercentDist = (n/total)*100) %>%
  filter(Day == "0") %>%
  arrange(PercentDist) %>%
  mutate(SampleID = factor(SampleID, levels = patient_order$SampleID))

df_1 %>%
  mutate(category = factor(category, levels = c("primitive", "GMP", "mature", "lymphoid", "erythroid", "pDC"))
         ) %>%
ggplot(. , aes(x = cell_status, y = PercentDist, fill = category)) +
  geom_col(aes (color = cell_status), position = "stack", width = 0.7, linewidth = 1.5) +
  geom_col(position = "stack", width = 0.6, linewidth = 1) +
  geom_text(
      aes(label = ifelse(PercentDist>2, paste0(round(PercentDist,1)), "" ) ),
      position = position_stack(vjust = 0.5), size = 3,color = "black" 
  ) +
  
  scale_color_manual(values = c("red", "royalblue")) +
  scale_fill_manual(
    values = c(primitive = "#ce9866", GMP = "#FDB462", 
               mature = "#B3DE69", lymphoid = "#BEBADA", 
               erythroid = "#F781BF", pDC = "#8DD3C7"), 
    labels = c("primitive\n(HSC, Prog)", "GMP", "mature\n(ProMono, Mono, cDC)",
               "lymphoid\n(ProB, B, Plasma, T, CTL, NK)", 
               "erythroid", "pDC") ) +
  scale_y_continuous(expand = c(0, 1)) +
  labs(y = "Percent Distribution (%)", x = "Patient Sample (Day 0)") +
  facet_wrap(~SampleID, ncol = 6, strip.position = "bottom", 
             labeller = labeller(SampleID = function(x) gsub("-D0", "", x)) ) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        legend.key.size = unit(0.3, "cm"), aspect.ratio = 2.5, 
        legend.position = "bottom", legend.direction = "vertical")

ggsave(paste0("plot/fig6/main_6c_cellpop_D0.png"), width = 7, height = 6, dpi = 600)
```

## fig 6d/suppfig 6b: inter-patient heterogenenity
```{r}
temp <- all_@meta.data %>%
  select(SampleID := orig.ident, PatientID, CellType, cell_status, Inflammation) %>%
  rownames_to_column("cell.barcode") %>%
  mutate(Day = gsub(".*-D", "", SampleID) ) %>%
  group_by(SampleID, Day) %>% # overall inflammation score per patient
  summarise(Inflammation = mean(Inflammation)) %>%
  filter(Day == "0") %>%
  inner_join(., df_1)
  
temp %>%
  # filter(category %in% c("primitive", "GMP", "mature")) %>% # main fig
  filter(!category %in% c("primitive", "GMP", "mature")) %>% # supp fig
  mutate(category = factor(category, levels = c("primitive", "GMP", "mature", "lymphoid", "erythroid", "pDC")),
         SampleID = factor(SampleID, levels = patient_order$SampleID)
         ) %>%
ggplot(., aes(x = PercentDist, y = Inflammation)) +
  geom_smooth(method = "lm", color = "black", se = F, size = 0.5) +
  stat_cor(method = "spearman", size = 3) +
  geom_point(aes(fill = SampleID), size = 2, alpha = .9, shape = 21, color = "grey10") +
  scale_fill_brewer(palette = "Paired") +
  theme_classic(base_size = 12) +
  labs(x="Percent Distribution (%)", y = "Mean Inflammatory Score") +
  facet_grid(category~cell_status, drop = T, axes = "all") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        aspect.ratio = 0.5, legend.key.size = unit(0.3, "cm"))

# ggsave(paste0("plot/fig6/main_6d_cellpop_inflam_main.png"), width = 7, height = 6, dpi = 600)
ggsave(paste0("plot/suppfig7/supp_7b_cellpop_inflam_supp.png"), width = 4.5, height = 6, dpi = 600)
```

# DEG: Differential between malignant and normal
- limiting to AML samples
- limiting to myeloid-monocytic lineage; excluding eryth
- refer to src/preprocess/6.scrna_mast.R
```{r}
selected <- subset(all_, subset = celltype %in% c("GMP", "Prog", "ProMono", "HSC", "cDC", "Mono")) 
```

## lineage specific: DEG between malignant and normal
- identifying genes/programs difference between the 2 differentiation stage

- Malignant Mature-like vs. Malignant Primitive-like
- Nonmalignant Mature-like vs. Nonmalignant Primitive-like
- Normal Mature-like vs. Normal Primitive-like

 
## tumor/normal specific: DEG between primitive and mature
- identifying genes/programs difference between the cell types in mature myeloid compartments ("ProMono", "Mono", "cDC")

- Mature-like Malignant vs. Mature-like Normal
- Mature-like Nonmalignant vs. Mature-like Normal
- Mature-like Nonmalignant vs. Mature-like Malignant


# Trajectory inference analysis

## slingshot
- limiting to AML samples
- limiting to myeloid-monocytic lineage; excluding eryth

- table(selected$seurat_clusters)
   0    1    2    3    4    5    6    7    8    9   10   11   12   13   14   15 
7421   33 4779 3979   10 1789   92 1297   41  921  352    7  247  256   35   89 
  16 
  59 
  - 1, 4, 6, 8, 11, 14, 15, 16 - removed for small cluster size (<100)
  - 10 was removed due to overlaps with erythroid pop; 12 removed due to distinctness from other HSC pop

(n=20442 after filtering)
   AML_malignant AML_nonmalignant   Healthy_normal 
           12826             3376             4240 
```{r}
table(selected$seurat_clusters)
table(selected$celltype, selected$seurat_clusters)

(DimPlot(selected, reduction = "umap", label = T, pt.size = .5, group.by = c("seurat_clusters")) + DimPlot(selected, reduction = "umap", label = T, pt.size = .5, group.by = c( "CellType"), cols = hierarchy_colors)) %>% set_alpha() & theme(aspect.ratio = 1, legend.position = "none")
# ggsave("plot/other/selected_bf_drop.png", width = 11, height = 5, dpi = 400)

## removing small clusters
clusters_to_remove <- c(1, 4, 6, 8, 10, 11, 12, 14, 15, 16)
selected <- subset(selected, subset = seurat_clusters %in% clusters_to_remove, invert = TRUE)
selected$seurat_clusters <- droplevels(selected$seurat_clusters)

(DimPlot(selected, reduction = "umap", label = T, pt.size = .5, group.by = c("seurat_clusters")) + DimPlot(selected, reduction = "umap", label = T, pt.size = .5, group.by = c( "CellType"), cols = hierarchy_colors)) %>% set_alpha() & theme(aspect.ratio = 1, legend.position = "none")
# ggsave("plot/other/selected_af_drop.png", width = 11, height = 5, dpi = 400)

saveRDS(selected, "output/rds/scrna_selected_obj.rds")
```

```{r}
pto <- slingshot(Embeddings(selected, "umap"),
                        clusterLabels = selected$seurat_clusters,
                        start.clus = 0, end.clus = c(7,13)) #


saveRDS(pto, "output/rds/scrna_selected_pseudotime_slingshot.rds")
```


## Trade-seq: Trajectory inference analysis

```{r}
counts_filt <- FetchData(selected, vars=VariableFeatures(selected), layer="counts")
counts_filt <- t(counts_filt)

BPPARAM <- BiocParallel::SnowParam(workers = 8, progressbar = TRUE)

sce <- fitGAM(counts = counts_filt, 
             sds = pto, 
             parallel=TRUE, 
             BPPARAM = BPPARAM,
             nknots = 6) # Reduced nknots slightly for speed if needed

saveRDS(sce, "output/rds/scrna_selected_pseudotime_fitGAM.rds")
```

- within lineage comparisons
  - associationTest: assesses whether average gene expression is associated with pseudotime - a particular lineage
  
- Between-lineage comparisons
  - patternTest: implements a statistical method that checks whether the smoothed gene expression is equal along pseudotime between two or multiple lineages.
  
```{r}
associationTest(sce, lineages = T) %>%
  rownames_to_column("feature_id") %>%
  arrange(-waldStat, -meanLogFC) %>%
  write.csv(., "output/supp/scrna/selected_avgGE_assoc_pseudotime.csv", row.names = F)

patternTest(sce, global = TRUE, pairwise = TRUE) %>%
  rownames_to_column("feature_id") %>% 
  filter(pvalue < 0.05, !is.na(pvalue)) %>% 
  arrange(-waldStat) %>%
  write.csv(., "output/supp/scrna/selected_patterntest.csv", row.names = F)
```

# plots & tbl

## load data
```{r}
load_pseudotime_data <- function() {
  all_selected <- readRDS("output/rds/scrna_selected_obj.rds")
  pto <- readRDS("output/rds/scrna_selected_pseudotime_slingshot.rds")
  sce <- readRDS("output/rds/scrna_selected_pseudotime_fitGAM.rds")
  cluster_name <- c(`0` = "HSC", `3` = "Prog", `9` = "GMP", `5` = "ProMono", `2` = "Mono", `7` = "cDC", `13` = "Mono")

  filt_counts <- FetchData(all_selected, vars = VariableFeatures(all_selected), layer="counts")
  sds <- SlingshotDataSet(pto)
  time <- slingAvgPseudotime(pto)

  df <- data.frame(slingReducedDim(sds),
                   cl = all_selected@meta.data$seurat_clusters,
                   pseudotime = time,
                   celltype = all_selected$CellType) %>%
    mutate(cluster = as.character(as.numeric(cl)))

  curves <- slingCurves(sds, as.df = TRUE)
  mst <- slingMST(sds, as.df = TRUE)
  mst$cluster_name <- cluster_name[mst$Cluster]

  all_selected$pseudotime <- time

  cluster_centers <- df %>%
    group_by(celltype) %>%
    summarize(umap_1 = median(umap_1), umap_2 = median(umap_2))

  return(list(
    all_selected = all_selected,
    pto = pto,
    sce = sce,
    filt_counts = filt_counts,
    sds = sds,
    time = time,
    df = df,
    curves = curves,
    mst = mst,
    cluster_name = cluster_name,
    cluster_centers = cluster_centers
  ))
}
```

  Lineage  cluster_name   
  <chr>    <chr>          
1 Lineage1 Lineage -> cDC 
2 Lineage2 Lineage -> Mono
3 Lineage3 Lineage -> GMP 

```{r}
pseudotime_list <- load_pseudotime_data()
list2env(pseudotime_list, envir = .GlobalEnv)
HVGs <- VariableFeatures(all_selected)

pseudotime_association <- read.csv("output/supp/scrna/selected_avgGE_assoc_pseudotime.csv") %>%
  filter(pvalue_1 < 0.01 | pvalue_2 < 0.01 | pvalue_3 < 0.01)

patternRes <- read.csv("output/supp/scrna/selected_patterntest.csv")
```

## tbl 19: overlapping DEG and pseudotime outcomes
```{r}
# 1) Read DEG table + add significance flag
read_deg <- function(path, p_adj = 0.05, lfc = 0.58) {
  read.csv(path) %>%
    mutate(sig = p_val_adj < p_adj & abs(avg_log2FC) > lfc) %>%
    arrange(avg_log2FC, p_val_adj)
}

# requires reference, HVGs, pseudotime_association, patternRes 
# 2) Build 3-way overlap + labels + "same direction" + common annotations
make_overlap_df <- function(a, b, c, set_map) {
  a <- read_deg(a)
  b <- read_deg(b)
  c <- read_deg(c)
  out <- full_join(a, b, by = "X", suffix = c("", ".1")) %>%
    full_join(c, by = "X", suffix = c("", ".2")) %>%
    filter(sig | sig.1 | sig.2) %>% # all must be non-NA
    mutate(
      s0 = replace_na(sig,   FALSE),
      s1 = replace_na(sig.1, FALSE),
      s2 = replace_na(sig.2, FALSE),
      pattern = paste0(as.integer(s0), as.integer(s1), as.integer(s2)),
      set = dplyr::recode(pattern, !!!set_map, .default = "NA"),
      # direction agreement (only compare where both contrasts exist)
      direction = sign(avg_log2FC),
      direction.1 = sign(avg_log2FC.1),
      direction.2 = sign(avg_log2FC.2),
      same = dplyr::case_when(
        pattern == "111" ~ direction == direction.1 & direction.1 == direction.2,
        pattern == "110" ~ direction == direction.1,
        pattern == "101" ~ direction == direction.2,
        pattern == "011" ~ direction.1 == direction.2,
        pattern %in% c("100", "010", "001") ~ TRUE,
        TRUE ~ NA
      ),
      pattern = paste0("s", pattern),
      is_inflam = X %in% reference$gene_id,
      is_variable = X %in% HVGs,
      is_pseudotime_ass = X %in% pseudotime_association$feature_id,
      is_diff_lineage_pattern = X %in% patternRes$feature_id
    ) %>%
    select(-s0, -s1, -s2) %>%
    filter(is_variable)
  out
}

```

### overlapping genes: malignant vs normal
- looking at genes that are contributing to "tumor" properties between malignant and normal (tumor intrinsic vs extrinsic)
- what genes cause malignant mature-like cells to be different from normal?
  - do they have inflammation genes certain persistant/absent programs

                   mature primitive
  AML_malignant      5906      6920
  AML_nonmalignant   2559       817
  Healthy_normal     1450      2790

- focusing on mature cells - understanding what are the different programs contributing to inflammation
- tumor intrinsic - malignant vs nonmalignant/normal (not included)
  - how does it affect the TME
- extrinsic - nonmalignant vs normal
  - what makes their inflammation different

      abnormal in AML                common    disease vs healthy 
                  312                    51                   145 
  genes affecting TME         TME extrinsic       tumor instrinic 
                   51                    78                   124 
tumor intrinsic (AML) 
                   63 

```{r}
df_malignant_normal <- make_overlap_df(
  a = "output/supp/scrna/findmarkers_mono_promono_cDC_malignant_normal.csv",
  b = "output/supp/scrna/findmarkers_mono_promono_cDC_nonmalignant_normal.csv",
  c = "output/supp/scrna/findmarkers_mono_promono_cDC_malignant_nonmalignant.csv",
  set_map = c(
    "111" = "common",
    "011" = "genes affecting TME",
    "110" = "abnormal in AML",
    "101" = "tumor instrinic",
    "010" = "TME extrinsic",
    "001" = "tumor intrinsic (AML)",
    "100" = "disease vs healthy")
)

table(df_malignant_normal$set)
write.csv(df_malignant_normal, "output/supp/scrna/deg_malignant_normal_pseudo.csv", row.names = F)
```

### overlapping genes: primitive vs mature
- looking at genes within malignant or normal cells that causes the activation of inflammation
  - for normal comparison - looks at the healthy template of differentiation switch
  - are the activation/causes of inflammation different - genes that are upregulated exclusively along the aberrant malignant trajectory

        abnormal in AML             common diff        diff lost in TME          malignant diff 
                    103                     445                     238                     420 
malignant specific diff             normal diff                TME diff 
                     49                     362                      14 

```{r}
df_differentiation <- make_overlap_df(
  a = "output/supp/scrna/findmarkers_normal_primitive_mature.csv",
  b = "output/supp/scrna/findmarkers_malignant_primitive_mature.csv",
  c = "output/supp/scrna/findmarkers_nonmalignant_primitive_mature.csv",
  set_map = c(
    "111" = "common diff",
    "011" = "abnormal in AML",
    "110" = "diff lost in TME",
    "101" = "malignant specific diff",
    "010" = "malignant diff",
    "001" = "TME diff",
    "100" = "normal diff")
)

write.csv(df_differentiation, "output/supp/scrna/deg_primitive_mature_pseudo.csv", row.names = F)
table(df_differentiation$set)
```

### saving overlap
- filtering for genes that are
  - sig diff between anyone of the comparisons between cell status in mature myeloid compartments
  - genes expressed in at least 10% in all groups
```{r}
df_differentiation_1 <- df_differentiation %>%
  filter(if_any(starts_with("pct"), ~ .x > 0.1) ) %>% 
  select(1:25) %>%
  left_join(df_malignant_normal %>% filter(if_any(starts_with("pct"), ~ .x > 0.1) ) , ., by="X", suffix = c(".mature", ".diff"))
  
write.csv(df_differentiation_1, paste0("output/scrna_deg_all.csv"), row.names = F)
```

## plot
## fig 6e: overall UMAP + pseudotime curves
```{r}
p <- DimPlot(object = all_selected, reduction = "umap", pt.size=.3, label = T) %>%
  set_alpha() + theme(aspect.ratio = 1, legend.position = "none")

q <- DimPlot(object = all_selected, reduction = "umap", group.by="CellType", shape.by = "cell_status", cols = hierarchy_colors, pt.size=.3, shuffle = T) %>%
  set_alpha() + 
  scale_shape_manual(values = c(17,16,15), name = NULL, 
                     labels = c("AML_malignant\n(n=12,826)", "AML_nonmalignant\n(n=3,376)", "Healthy_normal\n(n=4,240)")) +
  theme(aspect.ratio = 1, legend.key.size = unit(0.3, "cm")) + ggtitle("") +
  guides(shape = guide_legend(override.aes = list(size = 3)), 
         color = guide_legend(ncol=2, override.aes = list(size = 3)))

labelling <- mst %>%
  arrange(Order) %>%
  group_by(Lineage) %>%
  filter(row_number() == 1 | row_number() == n()) %>% ungroup() %>% distinct(umap_1, umap_2, cluster_name)

r <- FeaturePlot(object = all_selected, reduction = "umap", features = "pseudotime", pt.size=.3) %>%
  set_alpha() + 
  viridis::scale_color_viridis(name="Pseudotime", option = "C") +
  theme(aspect.ratio = 1, legend.position = "inside", 
        legend.position.inside = c(0.7, 0.2), legend.direction = "horizontal", 
        legend.key.width = unit(0.5, "cm"), legend.title.position = "top") + 
  ggtitle("") + 
  geom_point(data=mst %>% arrange(Order), aes(x=umap_1, y=umap_2), size = 3) +
  geom_path(data=mst %>% arrange(Order), aes(x=umap_1, y=umap_2, group=Lineage), lwd = 1) +
  ggrepel::geom_text_repel(data=labelling, aes(x=umap_1, y=umap_2, label=cluster_name), bg.colour = "white",  bg.r = 0.15, max.overlaps = Inf, nudge_x = -1.5)
  
p+q+r 

ggsave("plot/fig6/main_6e_selected_umap.png", dpi = 600, units = "in", width = 15, height = 4)
```

## fig 6e: transition of cell types along the lineage pseudotime

```{r}
pseudotime_df <- slingPseudotime(sds)
df <- FetchData(object = all_selected, vars = c("CellType", "celltype","cell_status"), clean = "none") %>%
  cbind(pseudotime_df)
colnames(df)[4:6] <- c("HSC \u2192 cDC", "HSC \u2192 Mono", "HSC \u2192 GMP")

temp <- df %>%
  pivot_longer(4:6, names_to = "Lineage", values_to = "pseudotime") %>%
  filter(!is.na(pseudotime)) %>%
  arrange(pseudotime) %>%
  mutate(Lineage = factor(Lineage, levels = rev(c("HSC \u2192 cDC", "HSC \u2192 Mono", "HSC \u2192 GMP"))),
         cell_status = factor(cell_status, levels = c("Healthy_normal", "AML_nonmalignant", "AML_malignant"))
         )

ggplot(temp, aes(x=pseudotime, y=cell_status, color=CellType, fill=CellType)) +
  geom_tile(alpha=0.5) +
  scale_color_manual(values = hierarchy_colors, guide = "none") +
  scale_fill_manual(values = hierarchy_colors) +
  facet_grid(Lineage ~ ., drop = T, scales = "free_y", axes = "all") +
  labs(y = "") +
  theme_classic() +
  theme(aspect.ratio = 0.1, strip.text.y = element_text(angle = 0), legend.key.size = unit(0.2, "cm"), text = element_text(size = 20)) + 
  guides(fill = guide_legend(ncol = 2, override.aes = list(alpha = 1))) 

ggsave("plot/fig6/main_6e_pseudotime_celltype.png", width = 12, height = 4, dpi = 600)
```

## fig 6f/suppfig 7c: transition of genes/signature along the pseudotime

- refer to function like plotSmoothers and GeneTrendCurve.Slingshot
```{r}
plot_pseudotime <- function(gene, file_prefix = NULL){
  labelling <- unique(mst[,c(4,6)]) %>%
    group_by(Lineage) %>%
    slice_tail(n = 1) %>% 
    mutate(Lineage = paste0("Lineage",Lineage),
           cluster_name = paste0("HSC \u2192 ",cluster_name))
  
  pseudotime_df <- slingPseudotime(sds)
  df <- FetchData(object = all_selected, vars = c(gene, "cell_status"), clean = "none") %>%
    rownames_to_column("cell") %>%
    cbind(pseudotime_df) %>%
    pivot_longer(gene, names_to = "Gene", values_to = "Expression") %>%
    pivot_longer(colnames(pseudotime_df), names_to = "Lineage", values_to = "Pseudotime") %>%
    left_join(labelling) %>%
    filter(!is.na(Pseudotime)) %>%
    mutate(cluster_name = factor(cluster_name, levels = c("HSC \u2192 GMP", "HSC \u2192 Mono", "HSC \u2192 cDC")))
  
  p <- ggplot(df, aes(x=Pseudotime, y=Expression, color=cell_status)) +
    geom_point(size=.01, alpha=.1) +
    geom_smooth(se = F, lwd = 1) + #"loess"
    scale_color_manual(values = color_celltype, name = "Status") +
    labs(title = gene) +
    theme_classic() + 
    theme(aspect.ratio = .75, text = element_text(size = 12),
          panel.grid.major = element_line(colour = "grey", linetype = 2),
          legend.position = "bottom", legend.direction = "vertical", 
          legend.text = element_text(size = 10),
          legend.key.size = unit(0.3, "cm")) +
    facet_wrap(~cluster_name, ncol = 1)
  
  if (!is.null(file_prefix)) {
    ggsave(plot = p, paste0(file_prefix, gene, ".png"), width = 4, height = 8, dpi = 600)
  } 
  return(p)
}
```

### selecting specific inflammation & differentiated related genes

- inflammatory related genes that are significantly different in malignant_vs_normal in mature myeloid compartment; that is also apparent in primitive_vs_mature

```{r}
## looking at trends for each gene individually
for (i in df_differentiation_1$X[df_differentiation_1$is_inflam]) {
  plot_pseudotime(i, file_prefix="plot/other/scrna_pseudo_")
}
```

- 43/61 was selected for association with immune/myeloid/aml related genes
- plotting only cdc and mono lineage; keeping gmp as supplement
```{r}
## heatmap
selected_gene <- list(
  "mature\nmyeloid" = c("CD14", "CD36", "CX3CR1", "MAFB", "MNDA", "FCGR3A", "FCN1",
                     "AZU1", "MPO", "CTSG", "ELANE", "SELL", "SRGN"),
  "immune\nregulation" = c("HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DRB1", "CD74", "CTSS", 
                        "FCER1G", "TYROBP", "LILRB2", "LILRA5"),
  "proinflammatory\nsignaling" = c("CXCL8", "CCL5", "SERPINA1", 
                                "S100A8", "S100A9", "S100A12", 
                                "C5AR1", "NCF2", "MAP3K8", "CFD"),
  "aml\ncore" = c("CDK6","HOXA9", "MYB"),
  "stress\nmetabolism\nother" = c("FOS", "HSPA5", "PMAIP1", "CLU", "MGST1", "CYP1B1")
)
temp <- names(selected_gene)
selected_gene <- stack(selected_gene)
colnames(selected_gene ) <- c("gene", "category")
selected_gene$category <- factor(selected_gene$category, levels = temp)

lineage_level <- c("HSC \u2192 cDC", "HSC \u2192 Mono", "HSC \u2192 GMP")
pseudotime_df <- slingPseudotime(sds) %>%
  as.data.frame()
colnames(pseudotime_df ) <- lineage_level

df <- FetchData(object = all_selected, vars = c(selected_gene$gene, "CellType", "celltype","cell_status"), clean = "none", layer = "scale.data") 

n = 1
df_1 <- df
df_1$Lineage <- pseudotime_df[,n]
df_1 <- df_1 %>% filter(!is.na(Lineage)) %>% arrange(Lineage)

ha1 <- columnAnnotation(
  cell_status = df_1$cell_status,
  celltype = df_1$CellType,
  col = list(
    cell_status = color_celltype,
    celltype = hierarchy_colors[c(1,2,5:16)]
  ),
  show_annotation_name = F
)

mat <- df_1[,selected_gene$gene] %>% as.matrix() %>% t()
col_fun <- circlize::colorRamp2(c(10, 5, 2.5, 1.83, 0.9, 0.45, 0, -2), c("#000004FF", "#420A68FF", "#932667FF", "#DD513AFF", "#FCA50AFF", "#FCFFA4FF", "lightyellow", "white"))

hm1 <- Heatmap(
  mat, top_annotation = ha1, col = col_fun, border_gp = gpar(col = "grey30"),
  show_row_dend = F, show_column_dend = F,
  show_column_names = F, cluster_columns = F, 
  cluster_rows = T, clustering_distance_rows = "euclidean", 
  column_split = df_1$cell_status, column_title = lineage_level[n],
  row_split = selected_gene$category, row_title_rot = 0, row_title_gp = gpar(fontsize = 11), 
  cluster_row_slices = F, show_row_names = F,
  name = "scaled", use_raster = T, raster_device = "png", raster_quality = 3,
  heatmap_legend_param = list(border = "black", labels_gp = gpar(fontsize = 10))
)

n = 2 # change to 3 for supp figure
df_1 <- df
df_1$Lineage <- pseudotime_df[,n]
df_1 <- df_1 %>% filter(!is.na(Lineage)) %>% arrange(Lineage)

ha1 <- columnAnnotation(
  cell_status = df_1$cell_status,
  celltype = df_1$CellType,
  col = list(
    cell_status = color_celltype,
    celltype = hierarchy_colors[c(1,2,5:16)]
  ),
  show_annotation_name = T
)

mat <- df_1[,selected_gene$gene] %>% as.matrix() %>% t()

hm2 <- Heatmap(
  mat, top_annotation = ha1, col = col_fun, border_gp = gpar(col = "grey30"),
  show_row_dend = F, show_column_dend = F,
  show_column_names = F, cluster_columns = F, 
  cluster_rows = T, clustering_distance_rows = "euclidean", 
  column_split = df_1$cell_status, column_title = lineage_level[n],
  row_split = selected_gene$category, row_title_rot = 0, row_title_gp = gpar(fontsize = 11), 
  cluster_row_slices = F, row_names_centered = T,
  name = "scaled", use_raster = T, raster_device = "png", raster_quality = 3,
  heatmap_legend_param = list(border = "black", labels_gp = gpar(fontsize = 10))
)
```

```{r}
png("plot/fig6/main_6e_hm_lineage_1_2.png", width = 9.5, height = 7.5, units = "in", res = 600)
draw(hm2 + hm1, auto_adjust = F, ht_gap = unit(-1, "mm"), merge_legends = T)

# png("plot/suppfig7/supp_7c_hm_lineage_3.png", width = 7, height = 7.5, units = "in", res = 600)
# draw(hm2, merge_legends = T)
```

# session info
```{r}
sessionInfo()
```

