---
title: "Integrating Proteomics BAML/PNNL"
description: "looking at DEG of genes detected in both proteomics and transcriptomics level"
date: "30 dec 25"
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: yes
---

# loading libraries
```{r message=FALSE, include=FALSE}
library(rstatix)
library(reshape2)
library(ggpubr)
library(ggrepel)
library(tidyverse)
library(ROTS)

library(ComplexHeatmap)
source("src/function/loading_data.R")
```

# integrating RNA and proteomics
- selecting the patients that have both data available
- using:
  RNA: VST transformed values 
  Proteomics: Corrected global gene proteomics from PNNL (TMT data processing)

<https://rdrr.io/bioc/GDCRNATools/src/R/gdcVoomNormalization.R>
<https://bioconductor.org/packages/devel/bioc/vignettes/ROTS/inst/doc/ROTS.pdf>
<https://academic.oup.com/nar/article/44/1/e1/2499676?login=false>

# loading data

- 111 samples

 1  2  3  4  5  6 
32 22 17 20 11  9 

- 60 - inflam_low; 51 inflam_high
```{r}
inflammatory_sig_genes <- readRDS("data/inflammatory_sig_genes.rds")

## sampleinfo
oshu_metadata <- read.csv("data/oshu/global_proteomics_metadata.csv") %>%
  mutate(id = gsub("PTRC_exp10_", "", Aliquot.Submitter.ID),
         clusters = cluster,
         cluster = ifelse(cluster %in% c(1,3,5), "inflam_low", "inflam_high")) %>%
  arrange(cluster, Sample_ID)
rownames(oshu_metadata) <- oshu_metadata$Sample_ID

## protein - 7084
global_proteomics <- read.csv("data/oshu/global_proteomics.csv", check.names = F, row.names = 1) %>%
  select(oshu_metadata$Sample_ID)

## transcript - 24751
expr_normalized <- readRDS("data/oshu/rnaseq_vst.rds") %>%
  slice_head(n=1, by=Gene) # duplicated genes are like SCARNA--; SNORA--; SNORD--; snoU--; U--; Y_RNA
gene_id <- setNames(expr_normalized$GeneID, expr_normalized$Gene)
rownames(expr_normalized) <- expr_normalized$Gene
expr_normalized <- expr_normalized[, oshu_metadata$Sample_ID]

## common genes - 6788
x <- intersect(rownames(global_proteomics), rownames(expr_normalized)) 
global_proteomics <- global_proteomics[x,]
expr_normalized <- expr_normalized[x,]
```

# tbl 22: correlation between transcript and protein abundance per gene
- 6788 common
```{r}
rna.data <- expr_normalized %>% 
  rownames_to_column("Gene") %>%
  reshape2::melt(variable.names = "Sample_ID", value.name = "GeneExp", id.vars = "Gene") %>%
  mutate(GeneID = gene_id[Gene])

prot.data <- global_proteomics %>% 
  rownames_to_column("Gene") %>%
  reshape2::melt(variable.names = "Sample_ID", value.name = "LogRatio", id.vars = "Gene")

corr_df <- inner_join(rna.data, prot.data) # 6788 common genes
corr_df <- corr_df %>%
  group_by(GeneID, Gene) %>%
  cor_test(vars = "GeneExp", vars2 = "LogRatio", method = "spearman")
write.csv(corr_df, "output/oshu_corr_global_prot_rna_per_gene.csv", row.names = F)
```

# running differential gene
```{r}
proteomics <- ROTS(data = global_proteomics, groups = oshu_metadata$cluster, B = 1000, K = 500, seed = 1234)

transcriptomics <- ROTS(data = expr_normalized, groups = oshu_metadata$cluster, B = 1000, K = 500, seed = 1234)

transcript <- data.frame(Gene = x,
                         ROTS_statistic = transcriptomics$d,
                         logfc = transcriptomics$logfc,
                         pvalue = transcriptomics$pvalue,
                         FDR = transcriptomics$FDR)

protein <- data.frame(Gene = x,
                      ROTS_statistic = proteomics$d,
                      logfc = proteomics$logfc,
                      pvalue = proteomics$pvalue,
                      FDR = proteomics$FDR)

## saving outcomes
write.csv(transcript, "output/supp/inflam_DEG_transcript.csv", row.names = F)
write.csv(protein, "output/supp/inflam_DEG_protein.csv", row.names = F)
```

# tbl 23: integrating the DEG of transcript and protein together
```{r}
corr_df <- read.csv("output/oshu_corr_global_prot_rna_per_gene.csv")
transcript <- read.csv("output/supp/inflam_DEG_transcript.csv")
protein <- read.csv("output/supp/inflam_DEG_protein.csv")

# Merge data on GeneID
data <- merge(transcript, protein, by = "Gene", suffixes = c("_rna", "_prot"))

quantile(data$logfc_prot)
quantile(data$logfc_rna)

thres_prot <- 0.58 # Threshold for proteomics 
thres_rna <- 1  # Threshold for transcriptomics

# Classify genes
data <- data %>%
  mutate(
    category = case_when(
      FDR_rna >= 0.05 | FDR_prot >= 0.05 ~ "C5.NS", # insignificant hits based on FDR
      logfc_rna >=  thres_rna & logfc_prot >=  thres_prot  ~ "C3.\u2191rna \u2191protein",
      logfc_rna <= -thres_rna & logfc_prot <= -thres_prot  ~ "C7.\u2193rna \u2193protein",
      logfc_rna >=  thres_rna & logfc_prot <= -thres_prot  ~ "C1.\u2191rna \u2193protein",
      logfc_rna >=  thres_rna & between(logfc_prot, -thres_prot, thres_prot) ~ "C2.\u2191rna NS protein",
      between(logfc_rna, -thres_rna, thres_rna) & logfc_prot <= -thres_prot  ~ "C4.NS rna \u2193protein",
      between(logfc_rna, -thres_rna, thres_rna) & logfc_prot >=  thres_prot  ~ "C6.NS rna \u2191protein",
      logfc_rna <= -thres_rna & between(logfc_prot, -thres_prot, thres_prot) ~ "C8.\u2193rna NS protein",
      logfc_rna <= -thres_rna & logfc_prot >=  thres_prot  ~ "C9.\u2193rna \u2191protein",
      .default = "C5.NS"
    )
  ) %>%
  separate_wider_delim(category, delim = ".", names = c("category_no", "category_name"), cols_remove = T)

data %>%
 mutate(sig = FDR_rna < 0.05 & FDR_prot < 0.05) %>%
 select(category_no, sig) %>%
 table()

```

## [not included in the table; replaced with elastic net regression] checking correlation of concordant genes C3 & C7

### concordance of 261 genes

- range of spearman correlation coefficient between transcript and proteins: 0.05 to 0.9 

```{r}
selected_gene <- data[data$category_no %in% c("C3", "C7"),]

hist(corr_df$cor[corr_df$Gene %in% selected_gene$Gene], breaks = 50)
# range(corr_df$cor[corr_df$Gene %in% selected_gene$Gene])

selected_gene <- selected_gene %>%
  filter(Gene %in% corr_df$Gene[corr_df$cor >= 0.5]) # 261->258
```

### correlation between protein abundance/ gene expression (of selected gene) with combination
```{r}
## overall correlation with proteomics
x <- as.data.frame(t(global_proteomics)) %>%
  select(selected_gene$Gene)
prot.data <- .run.cor(x, oshu_metadata[158:304], method = "spearman", min.n=5)
write.csv(prot.data, "output/supp/corr_prot_combi_all.csv", row.names = F)

## overall correlation with transcriptomics
x <- as.data.frame(t(expr_normalized)) %>%
  select(selected_gene$Gene)
rna.data <- .run.cor(x, oshu_metadata[158:304], method = "spearman", min.n=5)
write.csv(rna.data, "output/supp/corr_rna_combi_all.csv", row.names = F)

## selecting for significantly correlated genes
temp <- inner_join(rna.data[c(1,2,3,6,7,8,10,11)], 
                   prot.data[c(1,2,3,6,7,8,10,11)], 
                   by=c("Group", "Combination"), 
                   suffix = c(".R", ".P"))  %>%
  filter(p.adj.signif.R != "ns" & p.adj.signif.P != "ns") %>%
  mutate(in_inflam = Group %in% inflammatory_sig_genes$Gene)

write.csv(temp, "output/supp/corr_prot_rna_combi.csv", row.names = F)
```

## saving 
```{r}
# temp <- read.csv("output/supp/corr_prot_rna_combi.csv") # significant correlation w drug sens - 35

data <- data %>%
  group_by(category_no) %>%
  mutate(
    label = paste0("n=",n()),
    in_inflam = Gene %in% inflammatory_sig_genes$Gene, # inflam related: 171
    sig_corr_protrna = Gene %in% corr_df$Gene[corr_df$cor >= 0.5] # 1846
    # sig_corr_dss = Gene %in% temp$Group # sig corr w combi response - 154
         ) %>% 
  ungroup() %>%
  arrange(category_no, -abs(logfc_prot), -abs(logfc_rna))
  
write.csv(data, "output/oshu_rna_prot_overlap.csv", row.names = F, fileEncoding = "UTF-8")
```


# plots

## fig 7d: integrating rna and protein
```{r}
show <- c("SAMHD1", "NCF2", "CKAP4", "POU2F2", "CTSH", "KCTD12", "LRRC25", "CD14", "LILRA6")
```


```{r}
data <- read.csv("output/oshu_rna_prot_overlap.csv") %>%
  mutate(label2 = ifelse(sig_corr_dss & category_no == "C7", Gene, 
                         ifelse(((in_inflam & logfc_rna>2.1) | (Gene %in% show)) & category_no == "C3", Gene, NA)) ) %>%
  arrange(label2)

labels_ <- data[,c(10,12,11)] %>% 
  unique() %>% 
  as.data.frame() %>%
  arrange(category_no) 
labels_ <- rbind(c("C1", "n=0", "\u2191rna \u2193protein"), labels_)
labels_ <- rbind(labels_, c("C9", "n=0", "\u2193rna \u2191protein"))

labels_$x <- c(-2,0,2,-2,0,2,-2,0,2)
labels_$y <- c(5,5,5,0,0,0,-5,-5,-5)

data$category_name <- factor(data$category_name, levels = c(labels_$category_name))

table(data$category_no)

colors <- c("#377EB8","#984EA3","#E41A1C","grey","#4DAF4A","#A65628","gold","darksalmon", "#FF7F00")

# Plot
p <- ggplot(data, aes(x = logfc_prot, y = logfc_rna, colour = category_name, label= label2)) +
  geom_point(size = 2, alpha=.7) +
  scale_color_manual(values = colors, name = "Category") +
  geom_hline(yintercept = c(-1,1), linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = c(-0.58,0.58), linetype = "dashed", color = "gray70") +
  xlim(c(-2.1,2.2)) + ylim(c(-5,5)) +
  stat_cor(mapping = aes(x = logfc_prot, y = logfc_rna), method = "spearman", inherit.aes = F, label.y = -4) +
  geom_text_repel(min.segment.length = 0, box.padding=.5,color="black", 
                  size=3, max.overlaps = 30, force = 0.6) +
  labs(
    x = "Log2 Fold Change of Protein",
    y = "Log2 Fold Change of RNA"
  ) +
  theme_classic(base_size = 14, base_line_size = 0) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    aspect.ratio = 1,
    panel.border = element_rect(fill=NA, color = "black", linewidth = 1)
  ) +
  geom_text(data=labels_[c(3,7), ], mapping=aes(x,y,label=label), color="gray40") +
  guides(color = guide_legend(override.aes = list(size=3, alpha=1) ) ) +
  theme_pdf

ggsave(plot = p, "plot/fig7/main_7d_scatterplot_protrna.png", width = 7, height = 5, dpi = 600)
ggsave(plot = p, "plot/fig7/main_7d_scatterplot_protrna.pdf", width = 7, height = 5, bg = "transparent", device = cairo_pdf)
```


# session info
```{r}
sessionInfo()
```